---
title: "Drivers of variation"
author: "Urwah Nawaz"
date: "2026-01-30"
output:
  html_document:
    toc: true
    toc_float: 
      collapsed: true
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    autodep = TRUE,
    eval = TRUE,
	echo = TRUE,
	warning = FALSE,
	message = FALSE
)
```

# Introduction 


```{r set-up}
library(magrittr)
library(tibble)
library(reshape2)
library(corrplot)
library(dplyr)
library(ggvenn)
library(glue)
library(pander)
library(gridExtra)
library(variancePartition)
library(DT)
library(ggplot2)
library(UpSetR)
library(patchwork)
library(ggpubr)
library(variancePartition)
library(yaml)
library(EnsDb.Hsapiens.v86)
source(here::here("code/functions.R"))
```


```{r}
gene2symbol = genes(EnsDb.Hsapiens.v86, return.type="DataFrame") %>%
    as.data.frame()
gene2symbol %<>% 
    dplyr::select("EnsemblID" = "gene_id", gene_name)
```


```{r load-data}

paths <- yaml::read_yaml("config/paths-example.yaml")

expr_specs <- list(
  GTEx = list(dir = paths$bulk$formatted$gtex$dir,        file = "expression/GTEx-exp.csv"),
  PsychENCODE= list(dir = paths$bulk$formatted$psychencode$dir, file = "expression/PsychEncode-exp.csv"),
  BrainSpan= list(dir = paths$bulk$formatted$brainspan$dir,   file = "expression/BrainSpan-exp.csv"),
  BrainSeq = list(dir = paths$bulk$formatted$brainseq$dir,    file = "expression/BrainSeq-exp.csv"),
  HDBR = list(dir = paths$bulk$formatted$hdbr$dir,        file = "expression/HDBR-exp.csv")
)


e <- list()
e$Raw <- lapply(names(expr_specs), \(ds) {
  f <- file.path(expr_specs[[ds]]$dir, expr_specs[[ds]]$file)
  read_expr(f, ds)
})
names(e$Raw) <- names(expr_specs)
```

# Determining drivers of variation 

## Thresholding 
To investigate the impact of biological and technical factors on gene expression, we performed an analysis using the variancePartition package on the bulk RNA-seq data. Prior to the analysis, the count matrices were filtered to remove lowly expressed genes. The expression cut-off was selected at 1 RPKM/TPM in at least 10% of the samples 

```{r}
raw_list <- e$Raw

thresh_list <- lapply(raw_list, thresh)

# Summary table: genes before/after
gene_counts <- tibble(
  Dataset = names(raw_list),
  Genes_before = sapply(raw_list, nrow),
  Genes_after  = sapply(thresh_list, nrow)
) %>%
  mutate(Genes_removed = Genes_before - Genes_after,
         Pct_kept = round(100 * Genes_after / Genes_before, 1))
```

```{r}
DT::datatable(
  gene_counts,
  rownames = FALSE,
  options = list(
    pageLength = nrow(gene_counts),
    dom = "t"   
  ),
  caption = "Gene counts before and after thresholding - Adapted for Supplementary Table S5"
)
```


## Variance partition analysis 

variancePartition is a statistical and visualization framework which fits a linear mixed model for each gene and partitions the total variance into the contribution of each variable in the experimental design, plus the residual variance. The fractions sum to 1 for each gene, the variance fractions are easily interpretable across genes and sources of variation. 

Prior to performing a variancePartition analysis, the correlation of all variables considered for varianceParition analysis was performed using a Canonical Correlation Analysis (CCA), where variables with a highly correlating score (correlation > 0.95) were dropped, as high correlation can overestimate the contribution of variation. For each dataset, we performed variancePartition analysis with and without cell type composition estimates.  


```{r load-md}
md_specs <- list(
  GTEx = list(dir = paths$bulk$formatted$gtex$dir,        file = "metadata/GTEx-metadata.csv"),
  PsychENCODE= list(dir = paths$bulk$formatted$psychencode$dir, file = "metadata/PsychEncode-metadata.csv"),
  BrainSpan= list(dir = paths$bulk$formatted$brainspan$dir,   file = "metadata/BrainSpan-metadata.csv"),
  BrainSeq = list(dir = paths$bulk$formatted$brainseq$dir,    file = "metadata/BrainSeq-metadata.csv"),
  HDBR = list(dir = paths$bulk$formatted$hdbr$dir,        file = "metadata/HDBR-metadata.csv")
)

md <- lapply(names(md_specs), \(ds) {
  f <- file.path(md_specs[[ds]]$dir, md_specs[[ds]]$file)
  read.csv(f, check.names = FALSE)
})
names(md) <- names(md_specs)


```

### BrainSeq


* With cell type estimates  

```{r}
form.bseq <- ~ `Age (Numeric)` + (1|Regions) + (1|Sex) + RIN +  (1|Diagnosis) + `Mito Ratio` + `rRNA ratio` + `Total N Reads` + `Mapping Rate` +  (1|Period) + `Proportion of Neurons (MultiBrain)` + `Proportion of Astrocytes (MultiBrain)` + `Proportion of Microglia (MultiBrain)` + `Proportion of Endothelia (MultiBrain)`+ `Proportion of Microglia (MultiBrain)` + `Proportion of Oligodendrocytes (MultiBrain)` +`Total Assigned Gene`
C.bseq = canCorPairs(form.bseq, md$BrainSeq)
plotCorrMatrix( C.bseq )
```

```{r}
form <- ~`Age (Numeric)` + RIN +  `Mito Ratio` + `rRNA ratio` + `Total N Reads`  +  `Proportion of Neurons (MultiBrain)` +
  `Proportion of Astrocytes (MultiBrain)` + `Proportion of Microglia (MultiBrain)` +
  `Proportion of Endothelia (MultiBrain)`+ `Total Assigned Gene`
X <- model.matrix(form, data = md$BrainSeq)

apply(X, 2, sd)
apply(X, 2, var)

sd_vals <- apply(X, 2, sd)
sd_vals / min(sd_vals[sd_vals > 0])
```

- With cell type estimates 
```{r}

form.bseq <- ~ `Age (Numeric)` + (1|Regions) + (1|Sex) + RIN +  (1|Diagnosis) + `Mito Ratio` +
  `rRNA ratio` + `Total N Reads`  +  (1|Period) + `Proportion of Neurons (MultiBrain)` +
  `Proportion of Astrocytes (MultiBrain)` + `Proportion of Microglia (MultiBrain)` +
  `Proportion of Endothelia (MultiBrain)`+ `Total Assigned Gene`

md$BrainSeq %<>%
  column_to_rownames("SampleID")

num_cols <- sapply(md$BrainSeq, is.numeric)
md$BrainSeq[num_cols] <- scale(md$BrainSeq[num_cols])

varPar.ct.bseq <- fitExtractVarPartModel(thresh_list$BrainSeq, form.bseq, md$BrainSeq)
varPar.bseq = sortCols(varPar.bseq)
plot_varPart = plotVarPart(varPar.bseq)


save(varPar.bseq, file = here::here("output/variance-partition/results/BrainSeq-CT.Rda"))
```


- Without cell type estimates 

```{r}
form.bseq <- ~ `Age (Numeric)` + (1|Regions) + (1|Sex) + RIN +  (1|Diagnosis) + `Mito Ratio` + `rRNA ratio` + `Total N Reads` + `Mapping Rate` +  (1|Period) + +`Total Assigned Gene`
C.bseq = canCorPairs(form.bseq, md$BrainSeq)
plotCorrMatrix( C.bseq )
```

```{r}
form <- ~`Age (Numeric)` + RIN +  `Mito Ratio` + `rRNA ratio` + `Total N Reads`  + `Total Assigned Gene`
X <- model.matrix(form, data = md$BrainSeq)

apply(X, 2, sd)
apply(X, 2, var)

sd_vals <- apply(X, 2, sd)
sd_vals / min(sd_vals[sd_vals > 0])
```

```{r}

form.bseq <- ~ `Age (Numeric)` + (1|Regions) + (1|Sex) + RIN +  (1|Diagnosis) + `Mito Ratio` +
  `rRNA ratio` + `Total N Reads`  +  (1|Period) +  `Total Assigned Gene`

md$BrainSeq %<>%
  column_to_rownames("SampleID")

num_cols <- sapply(md$BrainSeq, is.numeric)
md$BrainSeq[num_cols] <- scale(md$BrainSeq[num_cols])

varPar.bseq <- fitExtractVarPartModel(thresh_list$BrainSeq, form.bseq, md$BrainSeq)
varPar.bseq = sortCols(varPar.bseq)
plot_varPart = plotVarPart(varPar.bseq)

plot_varPart

save(varPar.bseq, file = here::here("output/variance-partition/results/BrainSeq.Rda"))
```



### BrainSpan



```{r}
form.bspan <- ~`Age (Numeric)` + (1|Sex) + (1|Period) + (1|Regions) + RIN +  pH + `PMI (hours)` + `Dissection Score` + (1|DonorID)
C.bspan = canCorPairs(form.bspan, md$BrainSpan)
plotCorrMatrix( C.bspan )
```

```{r}
form.bspan <- ~`Age (Numeric)` + (1|Sex) + (1|Period) + (1|Regions) + RIN +  pH + `PMI (hours)` + `Dissection Score` + (1|DonorID)

varPar.bspan <- fitExtractVarPartModel(thresh_list$BrainSpan, form.bspan , md$BrainSpan)
varPar.bspan = sortCols(varPar.bspan )
plot_varPart = plotVarPart(varPar.bspan )

plot_varPart

save(varPar.bspan, file = here::here("output/variance-partition/results/BrainSpan.Rda"))
```


```{r}
form <- ~`Age (Numeric)` + RIN +  `Mito Ratio` + `rRNA ratio` + `Total N Reads`  + `Total Assigned Gene`
X <- model.matrix(form, data = md$BrainSeq)

apply(X, 2, sd)
apply(X, 2, var)

sd_vals <- apply(X, 2, sd)
sd_vals / min(sd_vals[sd_vals > 0])
```

```{r}

form.bseq <- ~ `Age (Numeric)` + (1|Regions) + (1|Sex) + RIN +  (1|Diagnosis) + `Mito Ratio` +
  `rRNA ratio` + `Total N Reads`  +  (1|Period) +  `Total Assigned Gene`

md$BrainSpan %<>%
  column_to_rownames("SampleID")

num_cols <- sapply(md$BrainSeq, is.numeric)
md$BrainSeq[num_cols] <- scale(md$BrainSeq[num_cols])

```


