---
title: "Drivers of variation"
author: "Urwah Nawaz"
date: "2026-01-30"
output:
  html_document:
    toc: true
    toc_float: 
      collapsed: true
    df_print: paged
    code_folding: hide
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    autodep = TRUE,
    eval = TRUE,
	echo = TRUE,
	warning = FALSE,
	message = FALSE
)
```

# Introduction 

This R markdown document contains the code to perform variance partition analysis on the bulk RNA-seq files for BITHub. All metadata and expression matrices were preprocessed using in the `notebooks/metadata-preprocess.Rmd`, whereas the deconvolution analysis for the cell type estimates were performed in `notebooks/bulk-deconvolution.Rmd`

In order to re-run the scripts, please change the location of input files. All directories are defined in `config/paths-example.yaml`. A copy of each processed metadata and deconvolution estimates are also available in `output/metadata/` and `output/deconvolution`, respectively.

```{r set-up}
library(magrittr)
library(tibble)
library(reshape2)
library(corrplot)
library(dplyr)
library(ggvenn)
library(glue)
library(pander)
library(gridExtra)
library(variancePartition)
library(DT)
library(ggplot2)
library(UpSetR)
library(patchwork)
library(ggpubr)
library(variancePartition)
library(yaml)
library(EnsDb.Hsapiens.v86)
source(here::here("code/functions.R"))
```


```{r}
gene2symbol = genes(EnsDb.Hsapiens.v86, return.type="DataFrame") %>%
    as.data.frame()
gene2symbol %<>% 
    dplyr::select("EnsemblID" = "gene_id", gene_name)
```


```{r load-data}

paths <- yaml::read_yaml(here::here("config/paths-example.yaml"))
expr_specs <- list(
  GTEx = list(dir = paths$bulk$formatted$gtex$dir,        file = "expression/GTEx-exp.csv"),
  PsychENCODE= list(dir = paths$bulk$formatted$psychencode$dir, file = "expression/PsychEncode-exp.csv"),
  BrainSpan= list(dir = paths$bulk$formatted$brainspan$dir,   file = "expression/BrainSpan-exp.csv"),
  BrainSeq = list(dir = paths$bulk$formatted$brainseq$dir,    file = "expression/BrainSeq-exp.csv"),
  HDBR = list(dir = paths$bulk$formatted$hdbr$dir,        file = "expression/HDBR-exp.csv")
)


e <- list()
e$Raw <- lapply(names(expr_specs), \(ds) {
  f <- file.path(expr_specs[[ds]]$dir, expr_specs[[ds]]$file)
  read_expr(f, ds)
})
names(e$Raw) <- names(expr_specs)
```

# Determining drivers of variation 

## Thresholding 
To investigate the impact of biological and technical factors on gene expression, we performed an analysis using the variancePartition package on the bulk RNA-seq data. Prior to the analysis, the count matrices were filtered to remove lowly expressed genes. The expression cut-off was selected at 1 RPKM/TPM in at least 10% of the samples 

```{r}
raw_list <- e$Raw

thresh_list <- lapply(raw_list, thresh)

# Summary table: genes before/after
gene_counts <- tibble(
  Dataset = names(raw_list),
  Genes_before = sapply(raw_list, nrow),
  Genes_after  = sapply(thresh_list, nrow)
) %>%
  mutate(Genes_removed = Genes_before - Genes_after,
         Pct_kept = round(100 * Genes_after / Genes_before, 1))
```

```{r}
DT::datatable(
  gene_counts,
  rownames = FALSE,
  options = list(
    pageLength = nrow(gene_counts),
    dom = "t"   
  ),
  caption = "Gene counts before and after thresholding - Adapted for Supplementary Table S5"
)
```


## Variance partition analysis 

variancePartition is a statistical and visualization framework which fits a linear mixed model for each gene and partitions the total variance into the contribution of each variable in the experimental design, plus the residual variance. The fractions sum to 1 for each gene, the variance fractions are easily interpretable across genes and sources of variation. 

Prior to performing a variancePartition analysis, the correlation of all variables considered for varianceParition analysis was performed using a Canonical Correlation Analysis (CCA), where variables with a highly correlating score (correlation > 0.95) were dropped, as high correlation can overestimate the contribution of variation. For each dataset, we performed variancePartition analysis with and without cell type composition estimates.  


```{r load-md}
md_specs <- list(
  GTEx = list(dir = paths$bulk$formatted$gtex$dir,        file = "metadata/GTEx-metadata.csv"),
  PsychENCODE= list(dir = paths$bulk$formatted$psychencode$dir, file = "metadata/PsychEncode-metadata.csv"),
  BrainSpan= list(dir = paths$bulk$formatted$brainspan$dir,   file = "metadata/BrainSpan-metadata.csv"),
  BrainSeq = list(dir = paths$bulk$formatted$brainseq$dir,    file = "metadata/BrainSeq-metadata.csv"),
  HDBR = list(dir = paths$bulk$formatted$hdbr$dir,        file = "metadata/HDBR-metadata.csv")
)

md <- lapply(names(md_specs), \(ds) {
  f <- file.path(md_specs[[ds]]$dir, md_specs[[ds]]$file)
  read.csv(f, check.names = FALSE)
})
names(md) <- names(md_specs)


```

### BrainSeq

- Without cell type estimates 

```{r}
form.bseq <- ~ `Age (Numeric)` + (1|Regions) + (1|Sex) + RIN +  (1|Diagnosis) + `Mito Ratio` + `rRNA ratio` + `Total N Reads` + `Mapping Rate` +  (1|Period) + +`Total Assigned Gene`
C.bseq = canCorPairs(form.bseq, md$BrainSeq)
plotCorrMatrix( C.bseq )
```

```{r}
form <- ~`Age (Numeric)` + RIN +  `Mito Ratio` + `rRNA ratio` + `Total N Reads`  + `Total Assigned Gene`
X <- model.matrix(form, data = md$BrainSeq)

apply(X, 2, sd)
apply(X, 2, var)

sd_vals <- apply(X, 2, sd)
sd_vals / min(sd_vals[sd_vals > 0])
```

```{r}

form.bseq <- ~ `Age (Numeric)` + (1|Regions) + (1|Sex) + RIN +  (1|Diagnosis) + `Mito Ratio` + `Mapping Rate` +
  `rRNA ratio` + `Total N Reads`  +  (1|Period) 

md$BrainSeq %<>%
  column_to_rownames("SampleID")

num_cols <- sapply(md$BrainSeq, is.numeric)
md$BrainSeq[num_cols] <- scale(md$BrainSeq[num_cols])

varPar.bseq_log <- fitExtractVarPartModel(log2(thresh_list$BrainSeq + 1), form.bseq, md$BrainSeq)
varPar.bseq_log = sortCols(varPar.bseq_log)
save(varPar.bseq_log, file = here::here("output/variance-partition/results/BrainSeq_log.Rda"))
```

```{r}

load(here::here("output/variance-partition/results/BrainSeq.Rda"))
load(here::here("output/variance-partition/results/BrainSeq_log.Rda"))


varPar.bseq_log %>% 
  as.data.frame() %>% 
  dplyr::select(Regions) %>%
  rownames_to_column("Gene") %>% 
  left_join(
  varPar.bseq %>%  as.data.frame() %>% 
  dplyr::select(Regions) %>%
  rownames_to_column("Gene"), by = "Gene")


plotVarPart(varPar.bseq) | plotVarPart(varPar.bseq_log)
```




* With cell type estimates  

```{r}


form.bseq <- ~ `Age (Numeric)` + (1|Regions) + (1|Sex) + RIN +  (1|Diagnosis) + `Mito Ratio` + `rRNA ratio` + `Total N Reads` + `Mapping Rate` +  (1|Period) + `Proportion of Neurons (MultiBrain)` + `Proportion of Astrocytes (MultiBrain)` + `Proportion of Microglia (MultiBrain)` + `Proportion of Endothelia (MultiBrain)` + `Proportion of Oligodendrocytes (MultiBrain)`
C.bseq = canCorPairs(form.bseq, md$BrainSeq)
plotCorrMatrix( C.bseq )
```

- With cell type estimates 
```{r}
form.bseq <- ~ `Age (Numeric)` + (1|Regions) + (1|Sex) + RIN +  (1|Diagnosis) + `Mito Ratio` + `rRNA ratio` + `Total N Reads` + `Mapping Rate` +  (1|Period) + `Proportion of Neurons (MultiBrain)` + `Proportion of Astrocytes (MultiBrain)` +  `Proportion of Endothelia (MultiBrain)` + `Proportion of Oligodendrocytes (MultiBrain)`

varPar.ct.bseq_log <- fitExtractVarPartModel(log2(thresh_list$BrainSeq + 1), form.bseq, md$BrainSeq)
varPar.ct.bseq_log = sortCols(varPar.ct.bseq_log )
save(varPar.ct.bseq_log, file = here::here("output/variance-partition/results/BrainSeq-CT_log.Rda"))
```

```{r fig.width=15, fig.height=5}
load(here::here("output/variance-partition/results/BrainSeq-CT.Rda"))

plotVarPart(varPar.ct.bseq_log) | plotVarPart(varPar.ct.bseq )
```


### BrainSpan



```{r}
form.bspan <- ~`Age (Numeric)` + (1|Sex) + (1|Period) + (1|Regions) + RIN +  pH + `PMI (hours)` + `Dissection Score` + (1|DonorID)
C.bspan = canCorPairs(form.bspan, md$BrainSpan)
plotCorrMatrix( C.bspan )
```

```{r}
num_cols <- sapply( md$BrainSpan, is.numeric)
md$BrainSpan[num_cols] <- scale( md$BrainSpan[num_cols])

md$BrainSpan %<>%
  column_to_rownames("SampleID")

varPar.bspan_log <- fitExtractVarPartModel(log2(thresh_list$BrainSpan + 1), form.bspan , md$BrainSpan)
varPar.bspan_log = sortCols(varPar.bspan_log)
save(varPar.bspan_log, file = here::here("output/variance-partition/results/BrainSpan_log.Rda"))
```


```{r}

load(here::here("output/variance-partition/results/BrainSpan_log.Rda"))

#write.csv(varPar.bspan_log, file = here::here("output/variance-partition/results/BrainSpan_varPart.csv"))
plot_varPart = plotVarPart(varPar.bspan )
plot_varPart

plotVarPart(varPar.bspan_log ) | plotVarPart(varPar.bspan )
```


```{r}
form.bspan <- ~`Age (Numeric)` + (1|Sex) + (1|Period) + (1|Regions) + RIN +  pH + `PMI (hours)` + `Dissection Score` + (1|DonorID) + `Proportion of Neurons (MultiBrain)` + `Proportion of Astrocytes (MultiBrain)` + `Proportion of Oligodendrocytes (MultiBrain)` + `Proportion of Endothelia (MultiBrain)`
C.bspan = canCorPairs(form.bspan, md$BrainSpan)
plotCorrMatrix( C.bspan )

varPar.bspan.CT_log <- fitExtractVarPartModel(log2(thresh_list$BrainSpan + 1), form.bspan , md$BrainSpan)





varPar.bspan.CT_log= sortCols(varPar.bspan.CT_log)
save(varPar.bspan.CT_log , file = here::here("output/variance-partition/results/BrainSpan-CT_log.Rda"))

```


```{r}
#load(here::here("output/variance-partition/results/BrainSpan-CT_log.Rda"))
write.csv(varPar.bspan.CT_log, file = here::here("output/variance-partition/results/BrainSpan_varPart_cellTypes.csv"))

plotVarPart(varPar.bspan_log ) | plotVarPart(varPar.bspan )
```

diagnostic 
```{r}
A <- varPar.bspan.CT_log
B <- read.csv(here::here("output/variance-partition/results/Gavin-comparison/Varpart_BrainSpan_CT_logged_gjs.csv"), row.names = 1, check.names = FALSE)

sum(rownames(B) %in% rownames(A))     
setdiff(rownames(B), rownames(A))    

g <- intersect(rownames(B), rownames(A))
k <- intersect(colnames(A), colnames(B))
A <- as.matrix(A[g, k]); B <- as.matrix(B[g, k])
D <- A - B

max(abs(D))                            
round(apply(abs(D), 2, max), 8)        
round(colMeans(D), 8)            

print(round(D, 6))
head(data.frame(max_diff = apply(abs(D), 1, max))[order(-apply(abs(D), 1, max)), , drop = FALSE], 20)

```


### GTEx 

#### Without cell type estimates

**Step 1 — Diagnostic: check collinearity among candidate covariates (full set)**
```{r}
# DIAGNOSTIC ONLY - canCorPairs check, full candidate covariate set
gtex.form <- ~ `Total N Reads` + `rRNA ratio` + (1|`Type of Batch`) + (1|`Age Interval`) + (1|Sex) + (1|Regions) + RIN + (1|`Hardy Scale`) + `PMI (hours)` + `Mapping Rate` + `Exonic Rate` + `Intronic Rate` + (1|`Autolysis Score`) + 
  `Mapped Unique (no dup flags)` + `End 2 Mismatch Rate` + `End 2 Antisense` + `End Sense 2` + 
  `End 1 Antisense` + `End 1 Sense` + `End 1 Sense percentage` + `Percentage Intragenic End 2 Reads`

C.gtex <- canCorPairs(gtex.form, md$GTEx)
plotCorrMatrix(C.gtex)
```

The canonical correlation analysis revealed several metadata had highly correlating scores. These variables were mostly sequencing metrics, including `End 1 Sense`, `End 2 Antisense`, `Mapped Unique (no dup flags)`, `End Sense 2`, and were dropped. In addition, `Autolysis Score` had a high number of missing values across datasets and was also dropped.

**Step 2 — Diagnostic: re-check collinearity after dropping correlated/incomplete variables**
```{r}
# DIAGNOSTIC ONLY - canCorPairs check, trimmed covariate set
gtex.form <- ~ `Total N Reads` + `rRNA ratio` + (1|`Type of Batch`) + (1|`Age Interval`) + (1|Sex) + (1|Regions) + RIN + (1|`Hardy Scale`) + `PMI (hours)` + `Mapping Rate` + `Exonic Rate` + (1|DonorID)

C.gtex <- canCorPairs(gtex.form, md$GTEx)
plotCorrMatrix(C.gtex)
```

**Step 3 — Diagnostic: check scale/variance of numeric covariates before scaling**
```{r}
# DIAGNOSTIC ONLY - numeric covariate scale check (not a fitted model)
form <- ~ `Total N Reads` + `rRNA ratio` + RIN + `PMI (hours)` + `Mapping Rate` + `Exonic Rate`
X <- model.matrix(form, data = md$GTEx)
apply(X, 2, sd)
apply(X, 2, var)
sd_vals <- apply(X, 2, sd)
sd_vals / min(sd_vals[sd_vals > 0])
```

**Step 4 — Scale numeric covariates**
```{r}
md$GTEx %<>%
  column_to_rownames("SampleID")
num_cols <- sapply(md$GTEx, is.numeric)
md$GTEx[num_cols] <- scale(md$GTEx[num_cols])
```

**Step 5 — FINAL MODEL (no cell types): fit and save**
```{r}
# FINAL MODEL - no cell types - this is the formula used to generate GTEx.Rda
gtex.form <- ~ `Total N Reads` + `rRNA ratio` + (1|`Type of Batch`) + (1|`Age Interval`) + (1|Sex) + (1|Regions) + RIN + (1|`Hardy Scale`) + `PMI (hours)` + `Mapping Rate` + `Exonic Rate` + (1|DonorID)

varPar.gtex_log <- fitExtractVarPartModel(log2(thresh_list$GTEx +1), gtex.form, md$GTEx)
varPar.gtex_log<- sortCols(varPar.gtex_log)
save(varPar.gtex_log, file = here::here("output/variance-partition/results/GTEx_log.Rda"))
```

**Step 6 — Load and plot final no-CT model**
```{r}
load(here::here("output/variance-partition/results/GTEx.Rda"))

plotVarPart(varPar.gtex_log) | plotVarPart(varPar.gtex)

```

#### With cell types

**Step 1 — Diagnostic: check collinearity among candidate covariates (full set + cell types)**
```{r}
# DIAGNOSTIC ONLY - canCorPairs check, full candidate covariate set + cell type proportions
gtex.form <- ~ `Total N Reads` + `rRNA ratio` + (1|`Type of Batch`) + (1|`Age Interval`) + (1|Sex) + (1|Regions) + RIN + (1|`Hardy Scale`) + `PMI (hours)` + `Mapping Rate` + `Exonic Rate` + `Intronic Rate` + (1|`Autolysis Score`) +
  `Mapped Unique (no dup flags)` + `End 2 Mismatch Rate` + `End 2 Antisense` + `End Sense 2` + 
  `End 1 Antisense` + `End 1 Sense` + `End 1 Sense percentage` + `Percentage Intragenic End 2 Reads` +
  `Proportion of Neurons (MultiBrain)` + `Proportion of Microglia (MultiBrain)` + `Proportion of Oligodendrocytes (MultiBrain)` + `Proportion of Endothelia (MultiBrain)` + `Proportion of Astrocytes (MultiBrain)`

C.gtex <- canCorPairs(gtex.form, md$GTEx)
plotCorrMatrix(C.gtex)
```

**Step 2 — FINAL MODEL (with cell types): fit and save**
```{r}
# FINAL MODEL - with cell types - this is the formula used to generate GTEx-CT.Rda
gtex.form <- ~ `Total N Reads` + `rRNA ratio` + (1|`Type of Batch`) + (1|`Age Interval`) + (1|Sex) + (1|Regions) + RIN + (1|`Hardy Scale`) + `PMI (hours)` + `Mapping Rate` +
  `Proportion of Neurons (MultiBrain)` + `Proportion of Oligodendrocytes (MultiBrain)` + `Proportion of Endothelia (MultiBrain)` + `Proportion of Astrocytes (MultiBrain)` + (1|DonorID)

varPar.gtex.ct_log <- fitExtractVarPartModel(log2(thresh_list$GTEx +1), gtex.form, md$GTEx)
varPar.gtex.ct_log <- sortCols(varPar.gtex.ct_log)
save(varPar.gtex.ct_log , file = here::here("output/variance-partition/results/GTEx-CT_log.Rda"))
```

**Step 3 — Load and plot final CT model**
```{r}
#load(here::here("output/variance-partition/results/GTEx-CT.Rda"))
#plotVarPart(varPar.gtex.ct) |  plotVarPart(varPar.gtex.ct_log)
```

<!-- ### GTEx -->

<!-- - Without cell type estimates  -->



<!-- ```{r} -->
<!-- gtex.form <- ~ `Total N Reads` + `rRNA ratio` + (1|`Type of Batch`)  + (1|`Age Interval`) + (1|Sex) + (1|Regions) + RIN + (1|`Hardy Scale`) + `PMI (hours)` + `Mapping Rate` + `Exonic Rate` + `Intronic Rate` + (1|`Autolysis Score`) +  -->
<!--   `Mapped Unique (no dup flags)` + `End 2 Mismatch Rate` +  `End 2 Antisense` +  `End Sense 2` +  -->
<!--   `End 1 Antisense` +  `End 1 Sense`  + `End 1 Sense percentage` +  `Percentage Intragenic End 2 Reads` -->
<!-- C.gtex = canCorPairs(gtex.form, md$GTEx) -->
<!-- plotCorrMatrix(C.gtex) -->
<!-- ``` -->

<!-- The canonical correlation analysis revealed several metadata had highly correlating score. These variables were mostly sequencing metrics, including End 1 Sense, End 2 Antisense, Mapped Unique (no dip flags), End Sense 2 and were dropped. In addition, Autolysis Score had a high number of missing values across datasets and was also dropped.  -->


<!-- ```{r} -->
<!-- gtex.form <- ~ `Total N Reads` + `rRNA ratio` + (1|`Type of Batch`)  + (1|`Age Interval`) + (1|Sex) + (1|Regions) + RIN + (1|`Hardy Scale`) + `PMI (hours)` + `Mapping Rate` + `Exonic Rate` + (1|DonorID) -->

<!-- C.gtex = canCorPairs(gtex.form, md$GTEx) -->
<!-- plotCorrMatrix(C.gtex) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- form <- ~`Total N Reads` + `rRNA ratio` + RIN + `PMI (hours)` + `Mapping Rate` + `Exonic Rate` -->

<!-- X <- model.matrix(form, data = md$GTEx) -->

<!-- apply(X, 2, sd) -->
<!-- apply(X, 2, var) -->

<!-- sd_vals <- apply(X, 2, sd) -->
<!-- sd_vals / min(sd_vals[sd_vals > 0]) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- md$GTEx %<>% -->
<!--   column_to_rownames("SampleID") -->

<!-- num_cols <- sapply(md$GTEx , is.numeric) -->
<!-- md$GTEx [num_cols] <- scale(md$GTEx[num_cols]) -->

<!-- #varPar.gtex <- fitExtractVarPartModel(thresh_list$GTEx, gtex.form  , md$GTEx) -->
<!-- #varPar.gtex  = sortCols(varPar.gtex  ) -->
<!-- #save(varPar.gtex , file = here::here("output/variance-partition/results/GTEx.Rda")) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- load( here::here("output/variance-partition/results/GTEx.Rda")) -->
<!-- plot_varPart = plotVarPart(varPar.gtex) -->
<!-- plot_varPart -->
<!-- ``` -->

<!-- - With cell types  -->

<!-- ```{r} -->
<!-- gtex.form <- ~ `Total N Reads` + `rRNA ratio` + (1|`Type of Batch`)  + (1|`Age Interval`) + (1|Sex) + (1|Regions) + RIN + (1|`Hardy Scale`) + `PMI (hours)` + `Mapping Rate` + `Exonic Rate` + `Intronic Rate` + (1|`Autolysis Score`) + -->
<!--   `Mapped Unique (no dup flags)` + `End 2 Mismatch Rate` +  `End 2 Antisense` +  `End Sense 2` +  -->
<!--   `End 1 Antisense` +  `End 1 Sense`  + `End 1 Sense percentage` +  `Percentage Intragenic End 2 Reads` + -->
<!--   `Proportion of Neurons (MultiBrain)` + `Proportion of Microglia (MultiBrain)` + `Proportion of Oligodendrocytes (MultiBrain)` +  `Proportion of Endothelia (MultiBrain)` +  `Proportion of Astrocytes (MultiBrain)`  -->
<!-- C.gtex = canCorPairs(gtex.form, md$GTEx) -->
<!-- plotCorrMatrix(C.gtex) -->

<!-- ``` -->

<!-- ```{r} -->
<!-- gtex.form <- ~ `Total N Reads` + `rRNA ratio` + (1|`Type of Batch`)  + (1|`Age Interval`) + (1|Sex) + (1|Regions) + RIN + (1|`Hardy Scale`) + `PMI (hours)` + `Mapping Rate` + -->
<!--   `Proportion of Neurons (MultiBrain)` +  `Proportion of Oligodendrocytes (MultiBrain)` +  `Proportion of Endothelia (MultiBrain)` +  `Proportion of Astrocytes (MultiBrain)` + (1|DonorID) -->



<!-- varPar.gtex.ct <- fitExtractVarPartModel(thresh_list$GTEx, gtex.form  , md$GTEx) -->
<!-- varPar.gtex.ct  = sortCols(varPar.gtex.ct  ) -->
<!-- save(varPar.gtex.ct  , file = here::here("output/variance-partition/results/GTEx-CT.Rda")) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- load(here::here("output/variance-partition/results/GTEx-CT.Rda")) -->
<!-- plot_varPart = plotVarPart(varPar.gtex.ct ) -->

<!-- plot_varPart -->
<!-- ``` -->

### HDBR 

```{r}
form.hdbr <- ~(1|Sex) + (1|Hemisphere) + (1|`Sequencing Batch`)+  `Total N Reads` + `Mapping Rate` + (1|Regions) + `PMI (hours)` + (1|DonorID)  + `Intronic Rate` + (1|`Age Interval`) 


C.hdbr = canCorPairs(form.hdbr, md$HDBR)
plotCorrMatrix(C.hdbr)
```

```{r}

md$HDBR %<>% 
   dplyr::select(-which(names(.) == "")) %>% 
 mutate(DonorID = factor(DonorID)) 

md$HDBR %<>% 
  distinct(SampleID, .keep_all=TRUE) %>% 
  column_to_rownames("SampleID")

num_cols <- sapply(md$HDBR , is.numeric)
md$HDBR[num_cols] <- scale(md$HDBR[num_cols])


varPar.hdbr_log <- fitExtractVarPartModel(log2(thresh_list$HDBR + 1), form.hdbr ,md$HDBR)
varPar.hdbr_log = sortCols(varPar.hdbr_log)
save(varPar.hdbr_log , file = here::here("output/variance-partition/results/HDBR-log.Rda"))
```


```{r}
load(here::here("output/variance-partition/results/HDBR-2.Rda"))
plotVarPart(varPar.hdbr_log) | plotVarPart(varPar.hdbr)

```


- With cell types 


```{r}
form.hdbr <- ~(1|Sex) + (1|Hemisphere) + (1|`Sequencing Batch`)+  `Total N Reads` + `Mapping Rate` + (1|Regions) + `PMI (hours)` + (1|DonorID)  + `Intronic Rate` + (1|`Age Interval`) + `Proportion of Neurons (MultiBrain)` + `Proportion of Astrocytes (MultiBrain)` + `Proportion of Oligodendrocytes (MultiBrain)` + `Proportion of Endothelia (MultiBrain)`


C.hdbr = canCorPairs(form.hdbr, md$HDBR)
plotCorrMatrix(C.hdbr)
```


```{r}
varPar.hdbr.ct_log <- fitExtractVarPartModel(log2(thresh_list$HDBR + 1), form.hdbr ,md$HDBR)
varPar.hdbr.ct_log = sortCols(varPar.hdbr.ct_log)
save(varPar.hdbr.ct_log , file = here::here("output/variance-partition/results/HDBR-CT-log.Rda"))
```

```{r}
load(here::here("output/variance-partition/results/HDBR-CT-2.Rda"))
plot_varPart = plotVarPart(varPar.hdbr)
plot_varPart
```


### PsychENCODE


- Without cell type 

```{r}
form.pe <- ~`Age (Numeric)` + (1|Sex) + (1|Period) + (1|`Cause of Death`) + (1|Diagnosis) + (1|Ethnicity) + (1|individualIDSource)

C.pe = canCorPairs(form.pe, md$PsychENCODE)
plotCorrMatrix(C.pe )

md$PsychENCODE %<>%
  dplyr::select(-which(names(.) == "")) %>% 
  distinct(SampleID, .keep_all = TRUE) %>%
  column_to_rownames("SampleID")

num_cols <- sapply(md$PsychENCODE, is.numeric)
md$PsychENCODE[num_cols] <- scale(md$PsychENCODE[num_cols])

varPar.pe_log <- fitExtractVarPartModel(log2(thresh_list$PsychENCODE +1 ), form.pe , md$PsychENCODE)
varPar.pe_log = sortCols(varPar.pe_log)
save(varPar.pe_log , file = here::here("output/variance-partition/results/PsychENCODE_log.Rda"))


```

```{r}

load(here::here("output/variance-partition/results/PsychENCODE.Rda"))
plot_varPart = plotVarPart(varPar.pe)
plot_varPart
```

- With cell type 

```{r}
form.pe <- ~`Age (Numeric)` + (1|Sex) + (1|Period) + (1|`Cause of Death`) + (1|Diagnosis) + (1|Ethnicity) + (1|individualIDSource) + `Proportion of Neurons (MultiBrain)` + `Proportion of Astrocytes (MultiBrain)` + `Proportion of Oligodendrocytes (MultiBrain)` + `Proportion of Endothelia (MultiBrain)`

C.pe = canCorPairs(form.pe, md$PsychENCODE)
plotCorrMatrix(C.pe )


varPar.pe.ct_log <- fitExtractVarPartModel(log2(thresh_list$PsychENCODE +1 ), form.pe , md$PsychENCODE)

varPar.pe.ct_log = sortCols(varPar.pe.ct_log)
save(varPar.pe.ct_log , file = here::here("output/variance-partition/results/PsychENCODE-CT_log.Rda"))
```


```{r}
load(here::here("output/variance-partition/results/PsychENCODE-CT.Rda"))
plot_varPart = plotVarPart(varPar.pe.ct)
plot_varPart
```
