---
title: "Pre-processing of metadata and expression matrices"
author: "Urwah Nawaz"
date: "2023-09-20"
output:
  html_document:
    toc: true
    toc_float: 
      collapsed: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    eval = TRUE,
	echo = TRUE,
	warning = FALSE,
	message = FALSE
)
```


# Introduction 

This R markdown document contains code to pre-process RNA-seq files for BITHub. Both expression matrices and metadata were retrieved as described in Table 1 in the GitHub Repository. 

In order to re-run the scripts, please change the location of input files. All directories are defined in `config/paths-example.yaml`. For each dataset, a pre-defined annotation file exists in the `data/annotations` folder. These files contains the original metadata name, and the BITHub column name for each dataset. 

A copy of each processed metadata will also be saved in 'output/metadata/'.

## Harmonization of metadata 
Developmental stages and anatomical regions were harmonised across datasets to ensure consistent filtering and downstream comparisons.

**Developmental stage definitions**:
Developmental stages were assigned using the lookup table in `code/def_stages.R`, which maps dataset-specific age fields to a standard BITHub stage label (e.g., prenatal vs postnatal groupings and broader developmental intervals). Where available, both the original age field and the harmonised stage label are retained in the final metadata.

**Anatomical region harmonisation**
Brain structures were mapped to harmonised anatomical regions using `code/def_regions.R`. Dataset-provided structure labels (or acronyms) were standardised and then grouped into broader regions (e.g., cortex, subcortex, cerebellum). Full definitions and mappings are provided in the Supplementary Table and referenced brain atlas documentation, enabling transparent interpretation of region-level comparisons.

## Selection of metadata variables 

Each metadata column was categorized into one of the following four categories:

* Phenotype attributes, describing observable characteristics relating to samples such as DonorID, Diagnosis, Sex and Age.

* Sample characteristics: referring to sample specific metrics such as site of sampling.

* Sequencing metrics: covering quality control (QC) metrics of a given RNA-seq sample, including QC output of bioinformatics protocol of a given sample.

* Cell type: capturing information on cell-types or cell specific metrics related to the single-nucleus RNA-seq or cell-type deconvolution for bulk RNA-seq datasets. 


Metadata were filtered prior to inclusion in BITHub to redunce redundancy and improve inerpretability. 

Filtering rules: 

* Drop variables with a single value across all samples 

* Drop variables with mostly missing values of `NA/Unknown`

* Drop categorial variables with > 50 categories (e.g DonorID)

* For sequencing metrics, drop variables that are highly correlated with another metric within the same dataset and keep one representative metric 


Metadata were filtered prior to inclusion in BITHub to reduce redundancy and improve interpretability. 

Filtering excluded:

* Variables with a single value across samples

* Variables with predominantly missing values

*  Categorical variables with more than 50 levels. 

Sequencing metrics that were highly correlated within a dataset were flagged to identify redundancy, and typically one representative metric was retained. Exceptions were made after manual review when keeping both metrics was useful and informative.


```{r libraries}
library(recount3)
library(magrittr)
library(tibble)
library(reshape2)
library(SummarizedExperiment)
library(corrplot)
library(dplyr)
library(ggvenn)
library(pander)
library(gridExtra)
library(variancePartition)
library(DT)
library(EnsDb.Hsapiens.v86)
library(AnnotationHub)
library(stargazer)
library(ggfortify)
library(patchwork)
library(foreach)
library(glue)
library(cowplot)
library(broom)
library(purrr)
library(glmpca)
library(stringr)
library(readxl)
library(readr)
library(DT)
library(naniar)
library(Seurat)
source(here::here("code/functions.R"))
source(here::here("code/def_stages.R"))
source(here::here("code/def_death.R"))
source(here::here("code/def_regions.R"))

paths <- yaml::read_yaml(here::here("config/paths-example.yaml"))
```



```{r set-up}
strc_acr <- map_df(structure_acronym, ~as.data.frame(.x), .id="id") %>% 
  set_colnames("Structure Acronym")
colnames(strc_acr)[2] = c("Structure")

rgns = map_df(regions, ~as.data.frame(.x), .id="id") %>% 
  set_colnames("Regions")
colnames(rgns)[2] = c("Structure Acronym")


rgns_ftl = map_df(regions_fetal, ~as.data.frame(.x), .id="id") %>% 
  set_colnames("Regions")
colnames(rgns_ftl)[2] = c("Structure Acronym")

order_vec <- c("Phenotype",
               "Sample characteristics",
               "Cell Type",
               "Sequencing metrics")
```

# Bulk RNA-seq data 

## BrainSeq

The BrainSeq data used for BITHub is obtained from BrainSeq Phase 2, which comprises RiboZero bulk RNA-seq data from the dorsolateral prefrontal cortex (DLPFC), and hippocampus (HIP). This cohort includes neurotypical individuals spanning from prenatal stages to ageing, as well as adult patients diagnosed with schizophrenia. 

The [gene expression matrices and the corresponding sample level metadata](https://s3.us-east-2.amazonaws.com/libd-brainseq2/rse_gene_unfiltered.Rdata) were downloaded from the BrainSeq consortium website. The release also includes pre-computed [cell type deconvolution estimates](https://s3.us-east-2.amazonaws.com/libd-brainseq2/methprop_pd.Rdata) which were generated as a part of the BrainSeq Phase 2 processing pipeline. 

### Metadata pre-processing

```{r load-bseq}
bseq.annot = read.csv(here::here("data/annotations/BrainSeq-metadata-annot.csv"), 
                      header = TRUE, 
                      check.names=FALSE)
load(file.path(paths$bulk$raw$brainseq, "rse_gene_unfiltered.Rdata"), envir = .GlobalEnv)
load(file.path(paths$bulk$raw$brainseq,"methprop_pd.Rdata"), envir = .GlobalEnv)
```
 
  
```{r preprocess-bseq}
x = rse_gene@colData %>% 
  as.data.frame() %>%
  t() %>% as.data.frame()
replicated <- colnames(x)[grep(",", x["SAMPLE_ID",])]
y <- as.list(x)
y[replicated] <- lapply(y[replicated], function(z) {
  # which variables to merge
  to.weight <- which(sapply(z, length) > 1 & sapply(z, class) %in% c("numeric", "integer"))
  # weighting of the merge
  weighting <- z$numReads # total reads
  weighting <- weighting / sum(weighting) # rather than a straight average, it's based on the number of reads
  
  # apply weighting
  z[to.weight] <- lapply(z[to.weight], function(zz) {
    if (length(weighting) == length(zz)) {
      sum(weighting * zz)
      } else {
          NaN
        }
        
      })
      
      # quickly fix character variables
      char <- which(sapply(z, length) > 1 & sapply(z, class) == "character")
      z[char] <- lapply(z[char], function(zz) {
        paste(zz, collapse = " & ")
      })
      
      return(z)
    })
    
w <- lapply(y, as.data.frame)
w <- do.call("rbind", w)
comp <- as.data.frame(pd)
comp <- comp[,57:64]
m <- match(rownames(comp), rownames(w)) # they are
md<- cbind(w, comp[m,])
    
md = rename_from_annot(md, bseq.annot)
    
    
# Adding features 
md %<>% 
  rownames_to_column("SampleID") %>%
  mutate(Period = ifelse(.$`Age (Numeric)` > 0, "Postnatal", "Prenatal"), 
                   `Structure Acronym` = gsub("HIPPO", "HIP", .$`Structure Acronym`),
                   Diagnosis = gsub("Schizo", "Schizophrenia", .$Diagnosis)) %>%
      mutate(Regions = add_feature(.$`Structure Acronym`, regions)) %>% 
      mutate(Age_rounded = as.character(sapply(na.omit(.$`Age (Numeric)`), num_to_round))) %>% as.data.frame() %>%
      mutate(`Age Interval` = as.character(add_feature(.$Age_rounded, age_intervals))) %>% 
      dplyr::select(-Age_rounded) %>%
      dplyr::select("SampleID", everything()) %>%
      as.data.frame()
```

After harmonization of metadata, a series of exploratory analyses, including correlation analysis was performed to remove redundant or non-informative metadata. 

For categorical metadata, a table is shown below which total number of unique values in each category.

```{r}
cat_tbl <- cat_summary(md, max_show = 50)


cat_tbl %>% as.data.frame() %>% DT::datatable(caption= "Summary of all categorial metadata in BrainSeq")
```


For sequencing metrics, a correlation analysis was performed using the `corrplot` R package. 
```{r}
seq_cols <- get_seq_metrics_cols(bseq.annot)  
seq_cols <- intersect(seq_cols, colnames(md))    
```

```{r}
res <- select_seq_metrics(
  md = md,
  seq_cols = seq_cols,
  corr_thresh = 0.95,
  missing_thresh = 0.5,
  min_unique = 2,
  prefer = "less_missing"
)

res$summary %>% DT::datatable(caption="Summary and correlation of all seqeuncing metrics metadata in BrainSeq")

```

```{r, fig.width=8, fig.height=8, fig.cap="Correlation plot showing correlation of sequencing metrics in BrainSeq"}

md.clean = md %>% 
  dplyr::select(contains(seq_cols))
md.clean = md.clean %>% select_if(~ !any(is.na(.)))
md.clean = md.clean[vapply(md.clean, function(x) length(unique(x)) > 1, logical(1L))]

M = cor(data.matrix(md.clean), use = "complete.obs")
corrplot(M, order='hclust',
         method = "circle", 
         number.cex = .50, 
         tl.cex=0.75)


```



Based on the above filtering criteria, followed by manual inspection, categorial metadata which have >50 unique or non-informative values were dropped. Additionally highly correlating sequencing metrics were also dropped. Additionally, FastQC metrics were also removed. 




```{r}
bseq.order = bseq.annot %>% 
  dplyr::mutate(`Include (Yes/No) - Interest` = ifelse(BITColumnName == "Reads Mapped to Mitochondrial Genes", "No", `Include (Yes/No) - Interest`)) %>%
  dplyr::mutate(`Include (Yes/No) - Interest` = ifelse(grepl("snp",BITColumnName),"Yes", `Include (Yes/No) - Interest`)) %>%
  dplyr::filter(`Include (Yes/No) - Interest` == "Yes") %>% 
  arrange(factor(Type, levels = order_vec)) %>% 
  arrange(factor(BITColumnName, levels = c("SampleID",
                                          "Age Interval",
                                          "Age (Numeric)",
                                          "Period",
                                          "DonorID", 
                                          "Sex", 
                                          "Ethnicity",
                                          "Diagnosis",
                                          "Structure Acronym",
                                          "Regions",
                                          "Proportion of Neurons (MultiBrain)",
                                          "Proportion of Astrocytes (MultiBrain)",
                                          "Proportion of Microglia (MultiBrain)",
                                          "Proportion of Oligodendrocytes (MultiBrain)",
                                          "Proportion of Endothelia (MultiBrain)"
                                          )))

```


```{r}
md %>% 
  dplyr::select(contains(bseq.order$BITColumnName)) %>%
  DT::datatable(caption = "BrainSeq metadata after harmonizing column names, ages and structure acronyms, and filtering for redundancy.")
md %<>% 
  dplyr::select(contains(bseq.order$BITColumnName))
```

In total BrainSeq contains `r length(unique(colnames(md)))` metadata columns. Here is the metadata summary:  


* Categorical values 

```{r}
res <- profile_df(md)
res$categorical_tables %>%
  do.call(rbind,.) %>% 
  DT::datatable(caption = "Summary for the selected categorial BrainSeq metadata, including number of samples per category and total proprotion of samples in the category")
```


```{r, fig.width = 10, fig.height = 14, fig.cap="Barplot showing counts of categorial metadata for BrainSeq"}
wrap_plots(res$categorical_plots, ncol = 3)
```


* Numerical values 

```{r}


res$numeric_summary %>% DT::datatable(caption = "Summary statistics for the selected numerical BrainSeq metadata, including number of samples per category, range of numerical distributions and number of samples with NAs.")
```


```{r, fig.width = 10, fig.height = 14, fig.cap = "Histogram showing distribution of numerical metadata."}
wrap_plots(res$numeric_plots, ncol = 3)
```





### Gene expression pre-processing
```{r exp-gene}
exp = rse_gene@assays@.xData$data$rpkm
rownames(exp) <- sub("\\.[0-9]*$", "", rownames(exp))
```
### Export data
```{r}
write.csv(exp, file = file.path(paths$bulk$formatted$brainseq$dir, paths$bulk$formatted$brainseq$expr))
write.csv(md, file = file.path(paths$bulk$formatted$brainseq$dir, paths$bulk$formatted$brainseq$metadata))
write.csv(bseq.order , file = file.path(here::here("data/annotations"), "BrainSeq-annot-final.csv"))
```

```{r include=FALSE}
rm(exp)
rm(md)
```

## BrainSpan

The BrainSpan dataset comprises 524 anatomically annotated post-mortem samples across multiple brain regions, obtained from donors spanning early postnatal development to ageing

The [expression matrix](http://www.brainspan.org/api/v2/well_known_file_download/267666525) and [metadata](http://www.brainspan.org/api/v2/well_known_file_download/267666525) were downloaded from the [BrainSpan Developmental Altas](https://www.brainspan.org/rnaseq/search/index.html). Additional metadata which includes sequencing metrics including PMI, RIN and pH was downloaded from the [Allen Brain Atlas](https://help.brain-map.org/download/attachments/3506181/Human_Brain_Seq_Stages.pdf?). 



### Metadata pre-processing

```{r load-bspan}
bspan.annot = read.csv(here::here("data/annotations/BrainSpan-metadata-annot.csv"), 
                       header = TRUE, 
                       check.names = FALSE)

columns.metadata = read.csv(file.path(paths$bulk$raw$brainspan, "columns_metadata.csv"), header = TRUE)
exp = read.csv(file.path(paths$bulk$raw$brainspan, "expression_matrix.csv"),
               header= FALSE, row.names= 1)
rows.metadata = read.csv(file.path(paths$bulk$raw$brainspan, "rows_metadata.csv"))
```


```{r md-bspan}
columns.metadata = rename_from_annot(columns.metadata, bspan.annot)

md = columns.metadata %>%
      mutate(Stage = add_feature(.$Age, stages),
             `Age Interval` = add_feature(.$Age, age_intervals),
             Age = gsub(" ","_", .$Age)) %>%
      mutate(SampleID = paste(DonorID, Age, `Structure Acronym`, Stage, sep = "_"),
             DonorName = gsub("\\.","_", .$DonorName),
            `Structure Acronym` = gsub("DFC", "DLPFC", `Structure Acronym`),
            `Structure Acronym` = gsub("MFC", "ACC", `Structure Acronym`)) %>%  ## Changed this for consistency across the datasets
  mutate(Regions = add_feature(.$`Structure Acronym`, regions)) %>%
    mutate(
    `Age (Numeric)` = case_when(
      str_detect(Age, regex("pcw", ignore_case = TRUE)) ~
        (as.numeric(str_remove(Age, "_pcw")) - 40) / 52,
      str_detect(Age, regex("mos", ignore_case = TRUE)) ~
        as.numeric(str_remove(Age, "_mos")) / 12,
      str_detect(Age, regex("yrs", ignore_case = TRUE)) ~
        as.numeric(str_remove(Age, "_yrs")),
      TRUE ~ NA_real_
    )
  ) %>% mutate(Period = ifelse(.$`Age (Numeric)` >= 0, "Postnatal", "Prenatal"),
                   colname = paste(DonorName, Age, `Structure Acronym`, sep = "_"))
```



```{r}
md.excel = read_excel(here::here("data/annotations/BrainSpan-additional.xlsx"),sheet =2, col_names = TRUE, skip =1) %>%
      as.data.frame() %>% mutate_at(.vars = "AllenInstituteID",
                                    .funs = gsub, pattern = "\\.", replacement = "\\_") %>%
      mutate_at(.vars = "Age", .funs = gsub, pattern = "PCW", replacement = "_pcw") %>%
      mutate_at(.vars = "Age", .funs = gsub, pattern = "M", replacement = "_mos") %>%
      mutate_at(.vars = "Age", .funs = gsub, pattern = "Y", replacement = "_yrs") %>%
      mutate_at(.vars = "Region/Area", .funs = gsub, pattern = "\\/", replacement = "-") %>%
      mutate(colname = paste(AllenInstituteID, Age, `Region/Area`, sep = "_")) %>%
      dplyr::select(-c(Agerange, Age, Description))

md = md %>%
      left_join(.,md.excel, by = "colname", keep = TRUE)
```

```{r}
md.excel = read_excel(here::here("data/annotations/BrainSpan-additional.xlsx"), sheet = 1, col_names = TRUE) %>%
      as.data.frame() %>%
      mutate_at(.vars="Internal ID", .funs = gsub, pattern = "\\.", replacement = "\\_") %>%
      dplyr::rename("Braincode" = "External ID")

md %<>%
  left_join(md.excel, by ="Braincode") %>%
      dplyr::select(-c("Age.y", "colname.y", "colname.x", "Gender", AllenInstituteID,
                     "Region/Area", "Internal ID")) %>%
      dplyr::rename("Ethnicity"="Ethn.")  %>%
  distinct(SampleID, .keep_all = TRUE)



md = md[!duplicated(md[,c('column_num')]),]
md %<>%
      dplyr::arrange(column_num)

md %<>%
  dplyr::select(-DonorID) %>% ## DonorID is numeric but DonorName is character, so replacing this
  dplyr::rename(DonorID = DonorName,
                `PMI (hours)` = "PMI",
                `Dissection Score` = Dissectionscore) %>%
  mutate(
  Ethnicity = recode(
    Ethnicity,
    "E"   = "CAUC",
    "H"   = "HISP",
    "As"  = "AS",
    "A"   = "AA",
    "A/E" = "Multiracial",
    .default = Ethnicity
  )
) %>% 
  mutate( `PMI (hours)` = as.numeric( `PMI (hours)`),
          `Dissection Score` = as.numeric(`Dissection Score`),
          `pH` = as.numeric(`pH`))
  
```

After harmonization of metadata, a series of exploratory analyses, including correlation analysis was performed to remove redundant or non-informative metadata. 

For categorical metadata, a table is shown below which total number of unique values in each category.


```{r}
cat_tbl <- cat_summary(md, max_show = 50)


cat_tbl %>% as.data.frame() %>% DT::datatable()
```

For sequencing metrics, a correlation analysis was performed using the `corrplot` R package. 

```{r}
seq_cols <- get_seq_metrics_cols(bspan.annot)  
seq_cols <- intersect(seq_cols, colnames(md))    
```

```{r}
res <- select_seq_metrics(
  md = md,
  seq_cols = seq_cols,
  corr_thresh = 0.95,
  missing_thresh = 0.5,
  min_unique = 2,
  prefer = "less_missing"
)

res$summary %>% DT::datatable()

```

```{r fig.cap="Correlation plot sequencing metrics in BrainSpan"}

md.clean = md %>% 
  dplyr::select(contains(seq_cols))
#md.clean = md.clean %>% select_if(~ !any(is.na(.)))
md.clean = md.clean[vapply(md.clean, function(x) length(unique(x)) > 1, logical(1L))]

M = cor(data.matrix(md.clean), use = "complete.obs")
corrplot(M, order='hclust',
         method = "circle", 
         number.cex = .50, 
         tl.cex=0.75)


```

Other than BrainCode, no metadata variables were removed from the BrainSpan meatdata. 

```{r}
bspan.order = bspan.annot %>%
  dplyr::filter(`Include (Yes/No) - Interest` == "Yes" | BITColumnName == "Dissection Score") %>%
  arrange(factor(Type, levels = order_vec)) %>%
  mutate(BITColumnName = ifelse(BITColumnName == "DonorName", "DonorID", BITColumnName)) %>%
  arrange(factor(BITColumnName, levels = c("Age Interval",
                                          "Age (Numeric)",
                                          "Period",
                                          "DonorID",
                                          "Sex",
                                          "Ethnicity",
                                          "Diagnosis",
                                          "Structure Acronym",
                                          "Regions",
                                          "SampleID",
                                          "Proportion of Neurons (MultiBrain)",
                                          "Proportion of Astrocytes (MultiBrain)",
                                          "Proportion of Microglia (MultiBrain)",
                                          "Proportion of Oligodendrocytes (MultiBrain)",
                                          "Proportion of Endothelia (MultiBrain)"
                                          )))

```


```{r bspan-md-final}
md %>%
  dplyr::select(contains(bspan.order$BITColumnName)) %>%
  DT::datatable(caption = "BrainSpan Metadata after harmonizing column names, ages and structure acronyms")

md %<>%
  dplyr::select(contains(bspan.order$BITColumnName)) %>%
  dplyr::select(SampleID, everything())
```


In total BrainSpan contains `r length(unique(colnames(md)))` metadata columns. Here is the metadata summary:  

* Categorical values 

```{r}
res <- profile_df(md)
res$categorical_tables %>%
  do.call(rbind,.) %>% 
  DT::datatable(caption = "Summary for the selected categorial BrainSpan metadata, including number of samples per category and total proprotion of samples in the category")
```


```{r, fig.width = 10, fig.height = 14, fig.cap="Barplot showing counts of categorial metadata for BrainSpan"}
wrap_plots(res$categorical_plots, ncol = 3)
```


* Numerical values 

```{r}


res$numeric_summary %>% DT::datatable(caption = "Summary statistics for the selected numerical BrainSpan metadata, including number of samples per category, range of numerical distributions and number of samples with NAs.")
```


```{r, fig.width = 10, fig.height = 14, fig.cap = "Histogram showing distribution of numerical metadata."}
wrap_plots(res$numeric_plots, ncol = 3)
```




### Expression data
```{r exp-rename}
colnames(exp) = md$SampleID
rownames(exp) <- rows.metadata$ensembl_gene_id
```



### Export

```{r}
write.csv(exp, file = file.path(paths$bulk$formatted$brainspan$dir, paths$bulk$formatted$brainspan$expr))
write.csv(md, file = file.path(paths$bulk$formatted$brainspan$dir, paths$bulk$formatted$brainspan$metadata))
write.csv(bspan.order, file = file.path(here::here("data/annotations"), "BrainSpan-annot-final.csv"))
```

```{r include =FALSE}
rm(exp)
rm(md)
rm(rows.metadata)
rm(columns.metadata)
```





## GTEx

The GTEx dataset provides RNA-seq profiles across a broad range of adult human tissues, including from several regions of the brain. 


Gene expression data and associated sample-level metadata were retrieved directly from the GTEx portal. For inclusion in BITHub, the dataset was subset to retain brain samples only, with non-brain tissues excluded. 

```{r}
gtex.annot = read.csv(here::here("data/annotations/GTEx-metadata-annot.csv"),
                      header = TRUE,
                      check.names= FALSE)

attributes = list.files(paths$bulk$raw$gtex, full.names = TRUE, pattern = "\\SampleAttributesDS.txt") # Sample attributes contains sample level information
phenotype = list.files(paths$bulk$raw$gtex, full.names = TRUE, pattern = "\\SubjectPhenotypesDS.txt") # Phenotype level information related to each donor
```


```{r}
md.attrbutes = read_tsv(attributes, col_types = c('.default' = 'c')) %>%
      dplyr::filter(SMTS == 'Brain') %>%
      mutate(SUBJID = sapply(str_split(SAMPID, pattern = "-"),
                             function(x) paste(x[1:2], collapse = '-'))) %>%
  left_join(read_tsv(phenotype, col_types = c('.default' = 'c')))  %>%
  as.data.frame()


md.attrbutes = rename_from_annot(md.attrbutes, gtex.annot)

md.attrbutes %<>%
  as.data.frame() %>%
  mutate(`Structure Acronym` = add_feature(.$`Structure Acronym`, structure_acronym)) %>%
      mutate(Regions = add_feature(.$`Structure Acronym`, regions),
             `Age Interval` = paste(.$`Age Interval`, "yrs", sep = ""),
             Sex = ifelse(Sex == 1, "M", "F"), ## Information taken from the corresponding excel sheet provided
             Period = "Postnatal") %>%
  as.data.frame() %>%
  mutate(
    `PMI (hours)` = as.numeric(`PMI (hours)`) / 60,

    `Hardy Scale` = recode(
      as.character(`Hardy Scale`), ## Coded from GTEx excel sheet
      "0" = "Ventilator Case",
      "1" = "Violent and fast death",
      "2" = "Fast death of natural causes",
      "3" = "Intermediate Death",
      "4" = "Natural Death",
      .default = as.character(`Hardy Scale`)
    ),

    `Autolysis Score` = recode(
      as.character(`Autolysis Score`), ## Coded from GTEx excel sheet
      "0" = "None",
      "1" = "Mild",
      "2" = "Moderate",
      "3" = "Severe",
      .default = as.character(`Autolysis Score`)
    )
  ) 
```


```{r gtex-exp}
exp = list.files(paths$bulk$raw$gtex, full.names = TRUE, pattern = "\\gene_tpm.gct.gz") # File used for expression matrix

exp = read.delim(exp, skip = 2)
colnames(exp) <- gsub("\\.", "-", colnames(exp)) # Changing expression file names to match metadata SampleIDs
exp %<>% column_to_rownames("Name")

exp = exp %>%
      dplyr::select(contains(md.attrbutes$SampleID))
message(paste0("Samples subsetted - Exp matrix contains ", ncol(exp), " samples"))
rownames(exp) <- sub("\\.[0-9]*$", "", rownames(exp))

md.attrbutes= md.attrbutes[which(md.attrbutes$SampleID %in% colnames(exp)),]  ## Only keeping sampleIDs which are present in the expression matrix
```


After harmonization of metadata, a series of exploratory analyses, including correlation analysis was performed to remove redundant or non-informative metadata. 

For categorical metadata, a table is shown below which total number of unique values in each category.

```{r}
cat_tbl <- cat_summary(md.attrbutes, max_show = 50)


cat_tbl %>% as.data.frame() %>% DT::datatable()
```


For sequencing metrics, a correlation analysis was performed for flagging potentially redundant variables.

```{r}
seq_cols <- get_seq_metrics_cols(gtex.annot)
seq_cols <- intersect(seq_cols, colnames(md.attrbutes))
```

```{r}
res <- select_seq_metrics(
  md = md.attrbutes,
  seq_cols = seq_cols,
  corr_thresh = 0.95,
  missing_thresh = 0.5,
  min_unique = 2,
  prefer = "less_missing"
)

res$summary %>% DT::datatable(caption = "Summary and correlation of all seqeuncing metrics metadata in GTEx")

```


```{r fig.cap="Correlation plot showing highly correlating metadata variables in GTEx"}

md.clean = md.attrbutes %>%
  dplyr::select(contains(seq_cols))
#md.clean = md.clean %>% select_if(~ !any(is.na(.)))
md.clean = md.clean[vapply(md.clean, function(x) length(unique(x)) > 1, logical(1L))]

M = cor(data.matrix(md.clean), use = "complete.obs")
corrplot(M, order='hclust',
         method = "circle",
         number.cex = .50,
         tl.cex=0.75)


```

Although several sequencing metrics were flagged for removal in the final metadata matrix, they were retained where they corresponded to standard quality control measures that provide relevant technical or experimental context.


```{r}
gtex.order = gtex.annot %>%
  dplyr::filter(`Include (Yes/No) - Interest` == "Yes") %>%
  mutate(Type = ifelse(Type == "Sample charactertics", "Sample characteristics", Type)) %>% ## In case there is any misspelling
  arrange(factor(Type, levels = order_vec)) %>%
  mutate(BITColumnName = ifelse(BITColumnName == "DonorName", "DonorID", BITColumnName)) %>%
  arrange(factor(BITColumnName, levels = c( "SampleID",
                                          "Age Interval",
                                          "Age (Numeric)",
                                          "Period",
                                          "DonorID",
                                          "Hardy Scale",
                                          "Sex",
                                          "Ethnicity",
                                          "Diagnosis",
                                          "Structure Acronym",
                                          "Regions",
                                          "Autolysis Score",
                                          "Sequencing Batch",
                                          "Type of Batch",
                                          "Date of Batch",
                                          "Genotype or Expression Batch ID",
                                          "Date of Genotype or Expression Batch",
                                          "Type of Genotype or Expression Batch",
                                          "BSS Collection side code",
                                          "Proportion of Neurons (MultiBrain)",
                                          "Proportion of Astrocytes (MultiBrain)",
                                          "Proportion of Microglia (MultiBrain)",
                                          "Proportion of Oligodendrocytes (MultiBrain)",
                                          "Proportion of Endothelia (MultiBrain)"
                                          )))
```

```{r gtex-md}
md.attrbutes %>%
  dplyr::select(contains(gtex.order$BITColumnName)) %>%
  DT::datatable(caption = "GTEx Metadata after harmonizing column names, ages and structure acronyms")

md.attrbutes %<>%
  dplyr::select(contains(gtex.order$BITColumnName)) %>%
dplyr::select(SampleID, everything())
```



In total the GTEx contains `r length(unique(colnames(md.attrbutes)))` metadata columns. Here is the metadata summary:  

* Categorical values 

```{r}
res <- profile_df(md.attrbutes)
res$categorical_tables %>%
  do.call(rbind,.) %>% 
  DT::datatable(caption = "Summary for the selected categorial GTEx metadata, including number of samples per category and total proprotion of samples in the category")
```


```{r, fig.width = 10, fig.height = 14, fig.cap="Barplot showing counts of categorial metadata for GTEx"}
wrap_plots(res$categorical_plots, ncol = 3)
```


* Numerical values 

```{r}

res$numeric_summary %>% DT::datatable(caption = "Summary statistics for the selected numerical GTEx metadata, including number of samples per category, range of numerical distributions and number of samples with NAs.")
```


```{r, fig.width = 10, fig.height = 5, fig.cap = "Histogram showing distribution of numerical metadata."}
wrap_plots(res$numeric_plots, ncol = 3)
```



### Export files
```{r}
write.csv(exp, file = file.path(paths$bulk$formatted$gtex$dir, paths$bulk$formatted$gtex$expr))
write.csv(md.attrbutes, file =file.path(paths$bulk$formatted$gtex$dir, paths$bulk$formatted$gtex$metadata))

write.csv(gtex.order, file = file.path(here::here("data/annotations"), "GTEx-annot-final.csv"))
```

```{r}
rm(gtex.order)
rm(exp)
rm(md.attrbutes)
```

## HDBR

The Human Developmental Biology Resource (HDBR) dataset comprises human brain RNA-seq samples collected across prenatal developmental stages from multiple anatomical sites.

The gene expression data and corresponding bioinformatics sequencing quality metrics were retrieved from the recount3 R package, which provides uniformly processed and normalized RNA-seq data derived from public repositories. 

Additional clinical and sample-level metadata, including developmental and experimental annotations not available in recount3, were obtained from the European Nucleotide Archive (ENA).

```{r hdbr-setup}
hdbr = recount3::create_rse_manual(
  project = "ERP016243",
  type = "gene"
)

hdbr.annot = read.csv(here::here("data/annotations/HDBR-metadata-annot.csv"),
                      header=TRUE,
                      check.names = FALSE)
hdbr_supp = read_excel(file.path(paths$bulk$raw$hdbr,"Amended Supplementary Table 1.xlsx"),
                       sheet = 1, skip =6) %>%
  as.data.frame()
hdbr_phenotype = read.csv(file.path(paths$bulk$raw$hdbr,"hdbr-phenotype.csv"),
                          header=TRUE)

colnames(hdbr_supp) = c("DonorID","Age", "SampleID", "Structure","Hemisphere", "Sex", "PMI")
hdbr_supp %<>% dplyr::select("DonorID","Structure", "PMI")
```


```{r}
md.hdbr = read.table(file.path(paths$bulk$raw$hdbr, "E-MTAB-4840-experiment-design.tsv"),
                     fill=TRUE, sep = "\t", header=TRUE )

```


```{r}
md.hdbr = md.hdbr[,1:15]

md.hdbr= rename_from_annot(md.hdbr, hdbr.annot)

md.hdbr %<>%
  as.data.frame() %>%
  dplyr::select(-c(Block, OntologyIndividual,  KaryotypeOntology, Organism, OrganismOntology,
                   StructureOntology, HemisphereOntology, OntologyAge)) %>%
  mutate(`Age Interval` = add_feature(.$Age, age_intervals))  %>%
  dplyr::rename("Structure"= "Structure Acronym") %>%
  left_join(strc_acr, by = "Structure") %>%
  left_join(rgns_ftl, by = "Structure Acronym") %>%
  mutate(Sex = toupper(Sex)) %>%
  mutate(Sex = ifelse(str_detect(Sex,"XX"), "F",
                      ifelse(str_detect(Sex, "XY"), "M", "Unknown" )),
         Period = c("Prenatal")) %>%
    mutate(Hemisphere = gsub("right", "R", Hemisphere),
         Hemisphere = gsub("left", "L", Hemisphere),
         Hemisphere = gsub("frontal", "Frontal", Hemisphere))

md = colData(hdbr) %>%
  as.data.frame() %>%
   dplyr::select(-contains("2"))
md = rename_from_annot(md, hdbr.annot)



hdbr_supp %<>%
  distinct(DonorID, .keep_all = TRUE) %>%
  mutate(PMI = ifelse(PMI == "UNKNOWN", NA, PMI))
md.hdbr %<>%
  left_join(hdbr_supp, by = c("DonorID"))
full.md.hdbr =left_join(md.hdbr, md)
full.md.hdbr %<>%
  mutate(PMI = as.numeric(PMI))

full.md.hdbr %<>%
  dplyr::filter(Structure.x != "stomach") %>%
  mutate(PMI = as.numeric(PMI) /60) %>%
  dplyr::rename(`PMI (hours)` = PMI) %>%
  mutate(DonorID = paste0(DonorID, "_donorID"))

```



```{r}
cat_tbl <- cat_summary(full.md.hdbr, max_show = 50)


cat_tbl %>% as.data.frame() %>% DT::datatable()
```

```{r}
seq_cols <- get_seq_metrics_cols(hdbr.annot)
seq_cols <- intersect(seq_cols, colnames(full.md.hdbr))
```

```{r}
res <- select_seq_metrics(
  md =full.md.hdbr,
  seq_cols = seq_cols,
  corr_thresh = 0.95,
  missing_thresh = 0.5,
  min_unique = 2,
  prefer = "less_missing"
)

res$summary %>% DT::datatable()

```





```{r fig.height=10, fig.width=10,fig.cap="Correlation plot showing highly correlating metadata variables in HDBR"}

md.clean = full.md.hdbr %>%
  dplyr::select(contains(seq_cols))
#md.clean = md.clean %>% select_if(~ !any(is.na(.)))
md.clean = md.clean[vapply(md.clean, function(x) length(unique(x)) > 1, logical(1L))]

M = cor(data.matrix(md.clean), use = "complete.obs")
corrplot(M, order='hclust',
         method = "circle",
         number.cex = .50,
         tl.cex=0.75)


```


```{r}
hdbr.order = hdbr.annot %>%
  dplyr::filter(`Include (Yes/No) - Interest` == "Yes") %>%
  mutate(Type = ifelse(Type == "Sample charactertics", "Sample characteristics", Type)) %>%
  arrange(factor(Type, levels = order_vec)) %>%
  mutate(BITColumnName = ifelse(BITColumnName == "DonorName", "DonorID", BITColumnName)) %>%
  arrange(factor(BITColumnName, levels = c("SampleID",
                                          "Age Interval",
                                          "Age (Numeric)",
                                          "Period",
                                          "DonorID",
                                          "Sex",
                                          "Ethnicity",
                                          "Diagnosis",
                                          "Structure Acronym",
                                          "Regions",
                                          "Hemisphere",
                                          "Proportion of Neurons (MultiBrain)",
                                          "Proportion of Astrocytes (MultiBrain)",
                                          "Proportion of Microglia (MultiBrain)",
                                          "Proportion of Oligodendrocytes (MultiBrain)",
                                          "Proportion of Endothelia (MultiBrain)",
                                          "PMI (hours)"
                                          )))
```



```{r hdbr-md}
full.md.hdbr %>%
  dplyr::select(contains(hdbr.order$BITColumnName)) %>%
  DT::datatable(caption = "HDBR Metadata after harmonizing column names, ages and structure acronyms")

full.md.hdbr %<>%
  dplyr::select(contains(hdbr.order$BITColumnName))
```


```{r hdbr-tpm}
assay(hdbr, "counts") = transform_counts(hdbr, round = TRUE)

hdbr.tpm = recount::getTPM(hdbr)

filter <- rowSums(hdbr.tpm, na.rm = TRUE) > 0.05
hdbr.filter.tpm = hdbr.tpm[filter,]
hdbr.filter.tpm <- thresh(hdbr.filter.tpm)


hdbr.final = hdbr.tpm %>%
  as.data.frame() %>%
  dplyr::select(contains(full.md.hdbr$SampleID)) %>%
  rownames_to_column("EnsemblID") %>%
  mutate_at(.vars = "EnsemblID", .funs = gsub, pattern = "\\.[0-9]*$", replacement = "")
```

### Export data
```{r}
write.csv(hdbr.final, file =file.path(paths$bulk$formatted$hdbr$dir,paths$bulk$formatted$hdbr$expr ))
write.csv(full.md.hdbr, file =file.path(paths$bulk$formatted$hdbr$dir,paths$bulk$formatted$hdbr$metadata))
write.csv(hdbr.order, file = file.path(here::here("data/annotations"), "HDBR-annot-final.csv"))
```

```{r}
rm(hdbr.final)
rm(full.md.hdbr)
rm(hdbr.order)
rm(hdbr)
rm(hdbr_phenotype)
```

## PsychENCODE

The PsychENCODE data used for BITHub were obtained from the PsychENCODE Capstone 4 release, a large-scale transcriptomic resource profiling post-mortem human brain tissue across neurotypical individuals and psychiatric disease cohorts. The dataset includes bulk RNA-sequencing data generated primarily from the DLPFC. 

The [gene expression matrix](http://resource.psychencode.org/Datasets/Derived/Controls_only_PEC_Gene_expression_matrix_normalized.txt) and corresponding [sample-level metadata](https://www.synapse.org/#!Synapse:syn18915911) were retrieved from the [PsychENCODE Capstone 4 project repository](http://resource.psychencode.org).

In addition, the PsychENCODE release includes pre-computed [cell type proportion estimates derived](http://resource.psychencode.org/Datasets/Derived/SC_Decomp/DER-24_Cell_fractions_Normalized.xlsx) from reference-based deconvolution approaches. These cell-type estimates were downloaded from the PsychENCODE Capstone 4 resource and incorporated into the BITHub preprocessing pipeline.


```{r pe-data}
pe.annot = read.csv(here::here("data/annotations/PsychEncode-metadata-annot.csv"),
                    header=TRUE,
                    check.names = FALSE) %>%
  dplyr::rename(OriginalMetadataColumnName = OriginalColumnName)  %>%
mutate(
     OriginalMetadataColumnName = gsub("\\.", "-", OriginalMetadataColumnName)
    )

md = list.files(paths$bulk$raw$psychencode, full.names = TRUE, pattern = "Job*") %>% read.csv(., header=TRUE) %>%
  dplyr::filter(individualID != "2015-1")
md.clinical = read.csv(file.path(paths$bulk$raw$psychencode, "PEC_capstone_data_map_clinical.csv"), header=TRUE, check.names = FALSE)

```

```{r}
exp = list.files(paths$bulk$raw$psychencode, full.names = TRUE, pattern = "\\Gene_expression_matrix_TPM.txt") %>%
      read.table(., header=TRUE, row.names = 1, check.names = FALSE)
```


```{r}
comp = list.files(paths$bulk$raw$psychencode, full.names = TRUE, pattern = "\\Cell_fractions*") %>% read_excel() %>%
      as.data.frame() %>%
      column_to_rownames("CellType")
```


A Venn diagram revealed column names that could not be matched to the metadata matrix. In total, 442 column names were unmappable and were therefore excluded; these columns corresponded to the GTEx dataset integrated into the PsychENCODE analysos. 

```{r}
library(VennDiagram)

pe.overlap = list("md individual ID" = md.clinical$`Synapse: individualID`,
                 #  "md pe individual ID" = md.clinical$`resource.psychencode.org: individualID`,
                  # "md original" = md$individualID,
                   "exp" = colnames(exp))
ggvenn(pe.overlap)

#x = calculate.overlap(pe.overlap)

#colnames(exp) %>%
 # as.data.frame() %>% dplyr::rename(SampleID = ".") %>%
  #dplyr::filter(!SampleID %in% md$individualID)
```

### Metadata pre-processing
```{r md-meta}

  # Fix existing columns

comp = as.data.frame(t(comp))

md %<>%
 left_join(comp %>% # So only those in the main MD can be used
  tibble::rownames_to_column("individualID"), by= "individualID")

md  = rename_from_annot(md, pe.annot, strict = FALSE)

# PCW to age numeric
md$`Age (Numeric)`[grepl("PCW", md$`Age (Numeric)`, ignore.case = TRUE)] = md$`Age (Numeric)`[grepl("PCW", md$`Age (Numeric)`)] %>%
  str_remove("PCW")%>%
  as.numeric() %>% `-` (40) %>% divide_by(52)

md$`Age (Numeric)` = gsub("90+", "91", md$`Age (Numeric)`)
md$`Age (Numeric)` =gsub("\\+", "", md$`Age (Numeric)`)


md$`Age (Numeric)` <- as.numeric(as.character(md$`Age (Numeric)`))

md %<>%
      dplyr::filter(Diagnosis == "Affective Disorder" |
               Diagnosis == "Autism Spectrum Disorder" |
               Diagnosis == "Bipolar Disorder" |
               Diagnosis == "Control" |
               Diagnosis == "Schizophrenia") %>%
      mutate(Structure = c("Dorsolateral Prefrontal Cortex"),  ## Adding name of structure
             `Structure Acronym` = c("DLPFC")) %>%
      mutate(Period = ifelse(.$`Age (Numeric)` >= 0, "Postnatal", "Prenatal"))  %>%
      mutate(Age_rounded = as.character(sapply(.$`Age (Numeric)`, num_to_round))) %>% as.data.frame() %>%
      mutate(`Age Interval` = as.character(add_feature(.$Age_rounded, age_intervals)),
             `Cause of Death` = as.character(add_feature(.$`Cause of Death`, death_cause))) %>%
      mutate(Regions = c("Cortex")) %>%
      mutate(DonorID = as.character(.$SampleID)) %>%
      dplyr::select(-Age_rounded) %>%
      as.data.frame() %>%
  mutate(Ethnicity = ifelse(Ethnicity == "HiSP", "HISP", Ethnicity))

exp = exp %>%
  dplyr::select(contains(md$SampleID))

md = md %>%
  dplyr::filter(SampleID %in% colnames(exp))
exp = exp[colnames(exp) %in% md$SampleID,]
md = md[md$SampleID %in% colnames(exp),]



```



As PsychENCODE did not contain sequencing metrics, no correlation analysis was performed. and an exploratory analysis of the categoiral metadata was performed to remove redundant data. All numerical metadata in this instance were kept. 


```{r}
cat_tbl <- cat_summary(md, max_show = 50)


cat_tbl %>% as.data.frame() %>% DT::datatable()
```

```{r}
pe.order = pe.annot %>%
  mutate(`Include (Yes/No) - Interest` = ifelse(BITColumnName == "individualIDSource","Yes", `Include (Yes/No) - Interest`),
         `Include (Yes/No) - Interest` = ifelse(BITColumnName == "DonorID","No", `Include (Yes/No) - Interest`)) %>% ## Amending this after discussion with Irina
  dplyr::filter(`Include (Yes/No) - Interest` == "Yes") %>%
  mutate(Type = ifelse(Type == "Sample charactertics", "Sample characteristics", Type)) %>%
  arrange(factor(Type, levels = order_vec)) %>%
  mutate(BITColumnName = ifelse(BITColumnName == "DonorName", "DonorID", BITColumnName)) %>%
  arrange(factor(BITColumnName, levels = c("SampleID",
                                           "Age Interval",
                                          "Age (Numeric)",
                                          "Period",
                                          "DonorID",
                                          "Sex",
                                          "Ethnicity",
                                          "Cause of Death",
                                          "Brain Weight",
                                          "Height",
                                          "Weight",
                                          "Diagnosis",
                                          "Structure Acronym",
                                          "Regions",
                                          "Capstone_4",
                                           "individualIDSource",
                                          "Proportion of Neurons (MultiBrain)",
                                          "Proportion of Astrocytes (MultiBrain)",
                                          "Proportion of Microglia (MultiBrain)",
                                          "Proportion of Oligodendrocytes (MultiBrain)",
                                          "Proportion of Endothelia (MultiBrain)"
                                          )))



```

```{r pe-md}
md %>%
   DT::datatable(caption = "PsychENCODE Metadata after harmonizing column names, ages and structure acronyms")

md %<>%
   dplyr::select(contains(pe.order$BITColumnName))
```

* Categorical values 

```{r}
res <- profile_df(md)
res$categorical_tables %>%
  do.call(rbind,.) %>% 
  DT::datatable(caption = "Summary for the selected categorial PsychENCODE metadata, including number of samples per category and total proprotion of samples in the category")
```


```{r, fig.width = 10, fig.height = 14, fig.cap="Barplot showing counts of categorial metadata for PsychENCODE"}
wrap_plots(res$categorical_plots, ncol = 3)
```


* Numerical values 

```{r}
res$numeric_summary %>% DT::datatable(caption = "Summary statistics for the selected numerical PsychENCODE metadata, including number of samples per category, range of numerical distributions and number of samples with NAs.")
```


```{r, fig.width = 10, fig.height = 14, fig.cap = "Histogram showing distribution of numerical metadata."}
wrap_plots(res$numeric_plots, ncol = 3)
```

 
### Export data

```{r}
write.csv(exp, file = file.path(paths$bulk$formatted$psychencode$dir, paths$bulk$formatted$psychencode$expr))
write.csv(md, file = file.path(paths$bulk$formatted$psychencode$dir,paths$bulk$formatted$psychencode$metadata), row.names = FALSE)

 write.csv(pe.order, file = file.path(here::here("data/annotations"), "PE-annot-final.csv"))
```


# Single nucleus RNA-seq data pre-process


## Human Cell Atlas (HCA)

Sn-RNAseq data were obtained from the Human Cell Atlas cortical cell-types resource. This dataset comprises transcriptomic profiles from 49,495 nuclei sampled across multiple regions of the human cerebral cortex, including the middle temporal gyrus (MTG), anterior cingulate cortex (ACC; ventral division of medial prefrontal cortex, A24), primary visual cortex (V1C), primary motor cortex (M1C), primary somatosensory cortex (S1C), and primary auditory cortex (A1C). Individual cortical layers were microdissected prior to profiling.

Expression matrices and accompanying metadata were downloaded directly from the Allen Brain Institute Cell Types Database (Human Multiple Cortical Areas, SMART-Seq) at:
https://brain-map.org/our-research/cell-types-taxonomies/cell-types-database-rna-seq-data/human-multiple-cortical-areas-smart-seq.

```{r}
exp <- read.csv(file.path(paths$single_nucleus$raw$hca, "matrix.csv"), check.names=FALSE)

meta <- read.csv(file.path(paths$single_nucleus$raw$hca, "metadata.csv"))
metaold <- read.csv(file.path(paths$single_nucleus$raw$hca, "metadata_release1.csv"))
hca.annot <- read.csv(here::here("data/annotations/HCA-annot.csv"),
                  header=TRUE,
                  check.names = FALSE) %>%
  dplyr::rename("OriginalMetadataColumnName" = "OriginalMetadata")

```

```{r}
rownames(exp) <- exp$sample_name
exp <- exp[,-1]
exp <- t(exp)

# create Seurat object 
HCA <- CreateSeuratObject(counts = exp, 
                             min.cells = round(ncol(exp) / 100), 
                             min.features = min.features, 
                             project = "CA") 

# read in metadata 
HCA$orig.celltype <-meta$cluster_label 


## Remove cells with no class 
keep <- which(!(HCA$orig.celltype == ""))
HCA <- subset(HCA, cells = keep) 

## Preprocess
max.depth <- get.max.depth(HCA)
HCA <- preprocess.fun(HCA, max.depth = max.depth) 

nCount = HCA$nCount_RNA %>%
      reshape2::melt() %>% set_colnames("nCount") %>%
      rownames_to_column("CellID")

nFeature = HCA$nFeature_RNA %>%
      reshape2::melt() %>% set_colnames("nFeature") %>%
      rownames_to_column("CellID")

mito_ratio = HCA$percent.mito%>%
      reshape2::melt() %>% set_colnames("Mito Ratio") %>%
      rownames_to_column("CellID")

## Save CPM 
dat <-  GetAssayData(object = HCA, assay = "RNA", slot = "counts") %>% as.data.frame()
dat <- make.cpm(dat) 
dat <- cbind(rownames(dat), dat) 
colnames(dat)[1] <- "Symbol" 
rownames(dat) <- 1:nrow(dat) 
```




```{r}
## Process metadata
# rename cell-types
rownames(meta) <- meta$sample_name
meta <- meta[colnames(HCA),]

meta %<>%
  mutate(
    MajorCelltype = case_when(
      str_detect(cluster_label, "Inh|GABA")        ~ "Inhibitory Neurons",
      str_detect(cluster_label, "Exc|Glut")        ~ "Excitatory Neurons",
      str_detect(cluster_label, "Astro")            ~ "Astrocytes",
      str_detect(cluster_label, "Endo|VLMC|Peri")  ~ "Vasculature",
      str_detect(cluster_label, "Micro")            ~ "Microglia",
      str_detect(cluster_label, "Oligo")            ~ "Oligodendrocytes",
      str_detect(cluster_label, "OPC")              ~ "OPCs",
      str_detect(cluster_label, "Non")              ~ "Unassigned Nonneuronal",
      TRUE                                          ~ NA_character_
    )
  )


# add further donor information from the metadata of HCA release 1

m <- match(meta$external_donor_name_label, metaold$donor)
meta$donor_age_days <- as.numeric(as.character(metaold$age_days[m]))
meta$`Age (Numeric)` <- meta$donor_age_days / 365
meta %<>% mutate(Age_rounded = as.character(sapply(na.omit(.$`Age (Numeric)`), num_to_round))) %>%
    mutate(`Age Interval` = add_feature(.$Age_rounded, age_intervals)) %>%
    dplyr::select(-Age_rounded) %>%
    mutate(Regions = add_feature(.$region_label, regions)) 


meta  = rename_from_annot(meta , hca.annot, strict = FALSE)


meta %<>% left_join(nCount, by = "CellID") %>%
      left_join(nFeature, by = "CellID") %>%
  left_join(mito_ratio, by="CellID")

meta %<>% 
  mutate(
    `Structure Acronym` = dplyr::recode(
      `Structure Acronym`,
      "M1lm" = "M1C",
      "M1ul" = "M1C",
      "S1lm" = "S1C",
      "S1ul" = "S1C",
      .default = `Structure Acronym`
    )
  )



 
```



```{r}
cat_tbl <- cat_summary(meta , max_show = 50)


cat_tbl %>% as.data.frame() %>% DT::datatable()
```

```{r}
seq_cols <- get_seq_metrics_cols(hca.annot)
seq_cols <- intersect(seq_cols, colnames(meta))
```

```{r}
res <- select_seq_metrics(
  md = meta,
  seq_cols = seq_cols,
  corr_thresh = 0.9,
  missing_thresh = 0.5,
  min_unique = 2,
  prefer = "less_missing"
)
res$summary %>% DT::datatable()

```

```{r}
meta %<>% 
  dplyr::select(contains(hca.annot %>% 
                           dplyr::filter(`Include (Yes/No) - Interest` == "Yes") %>% 
                           with(BITColumnName))) %>% 
  dplyr::select(-c(donor_sex_color, donor_sex_order,subclass_color, subclass_order ))
```

In total the HCA dataset contains `r length(unique(colnames(meta)))` metadata columns. Here is the metadata summary:  

```{r}
res <- profile_df(meta)

res$numeric_summary %>% DT::datatable()
```


```{r}
res$categorical_tables %>%
  do.call(rbind,.) %>%
  DT::datatable()
```

```{r, fig.width = 10, fig.height = 14}
wrap_plots(res$categorical_plots, ncol = 3)
```

```{r, fig.width = 10, fig.height = 14}
wrap_plots(res$numeric_plots, ncol = 3)
```



```{r}
write.csv(meta, file = file.path(paths$single_nucleus$formatted$hca$dir, paths$single_nucleus$formatted$hca$metadata), quote = FALSE, row.names = FALSE)
#write.csv(names, file= "../annotations/HCA-annot2.csv")

write.csv(dat, file = file.path(paths$single_nucleus$formatted$hca$dir, paths$single_nucleus$formatted$hca$expr), row.names = TRUE, col.names = TRUE, quote = FALSE)
```


## Velmeshev et al

Sn-RNAseq data were obtained from the study by Velmeshev et al. investigating cell-type–specific transcriptional changes in autism spectrum disorder (ASD). Post-mortem human cortical tissue samples from ASD patients and neurotypical control subjects were processed for nuclei isolation and profiled using the 10× Genomics single-nucleus RNA-sequencing platform.

Processed expression matrices and associated metadata were downloaded from the UCSC Autism Cell Atlas portal at:
https://autism.cells.ucsc.edu/.


```{r}
vel = file.path(paths$single_nucleus$raw$velmeshev)
vel.annot = read.csv(here::here("data/annotations/Velmeshev-annot.csv"),
                     header=TRUE,
                     check.names = FALSE)
## Load

dat <- Read10X(vel)

VL <- CreateSeuratObject(counts = dat,
                             min.cells = round(ncol(dat) / 100),
                             min.features = min.features,
                             project = "VL")
rm(dat)

## Annotate
meta <- read.table(file.path(vel, "meta.txt"), sep = "\t", header = TRUE)
m <- match(colnames(VL), meta$cell)

VL$orig.celltype <- meta$cluster[m]
VL$Region <- meta$region[m]
VL$Disorder <- meta$diagnosis[m]
VL$orig.ident <- paste0("Vel_", substr(meta$cell[m], start = 1, stop = 16))
VL$Individual <- meta$individual[m]

## Preprocess and normalise
max.depth <- get.max.depth(VL)
VL <- preprocess.fun(VL, max.depth = max.depth)



## CPM
dat <- GetAssayData(object = VL, assay = "RNA", slot = "counts") %>% as.data.frame()
dat <- make.cpm(dat)
dat <- cbind(rownames(dat), dat)
colnames(dat)[1] <- "Symbol"
rownames(dat) <- 1:nrow(dat)

```


```{r}

meta <- meta %>%
  mutate(
    MajorCelltype = case_when(
      str_detect(cluster, "IN")        ~ "Inhibitory Neurons",
      str_detect(cluster, "^L")        ~ "Excitatory Neurons",
      str_detect(cluster, "AST")       ~ "Astrocytes",
      str_detect(cluster, "Endo")      ~ "Endothelia",
      str_detect(cluster, "Micro")     ~ "Microglia",
      str_detect(cluster, "Oligo")     ~ "Oligodendrocytes",
      str_detect(cluster, "OPC")       ~ "OPCs",
      str_detect(cluster, "NRGN")      ~ "NRGN Neurons",
      str_detect(cluster, "mat")       ~ "Maturing Neurons",
      TRUE                             ~ NA_character_
    )
  ) %>%
  mutate(
    # only add Age_rounded for non-NA age, keep NA otherwise
    Age_rounded = if_else(
      is.na(age),
      NA_character_,
      as.character(sapply(age, num_to_round))
    ),
    `Age Interval` = add_feature(Age_rounded, age_intervals),
    region = str_replace(region, "PFC", "DLPFC"), ## change PFC to DLPFC
    Regions = add_feature(region, regions)
  ) %>%
  dplyr::select(-Age_rounded) %>%
  mutate(individual = paste0(individual, "_donorID"))


# save
m <- match(colnames(VL), meta$cell)
meta <-  meta[m,]

meta = rename_from_annot(meta, vel.annot, strict=FALSE)
meta %>% DT::datatable()

```



After harmonization of metadata, a series of exploratory analyses, including correlation analysis was performed to remove redundant or non-informative metadata. 

For categorical metadata, a table is shown below which total number of unique values in each category.

```{r}
cat_tbl <- cat_summary(meta , max_show = 50)


cat_tbl %>% as.data.frame() %>% DT::datatable()
```


```{r}
seq_cols <- get_seq_metrics_cols(vel.annot)
seq_cols <- intersect(seq_cols, colnames(meta))
```

```{r}
res <- select_seq_metrics(
  md = meta,
  seq_cols = seq_cols,
  corr_thresh = 0.9,
  missing_thresh = 0.5,
  min_unique = 2,
  prefer = "less_missing"
)
res$summary %>% DT::datatable()

```

```{r fig.cap="Correlation plot showing of correlation of sequencing metrics in Velmeshev et al"}
md.clean = meta %>%
  dplyr::select(contains(seq_cols))
#md.clean = md.clean %>% select_if(~ !any(is.na(.)))
md.clean = md.clean[vapply(md.clean, function(x) length(unique(x)) > 1, logical(1L))]

M = cor(data.matrix(md.clean), use = "complete.obs")
corrplot(M, order='hclust',
         method = "circle",
         number.cex = .50,
         tl.cex=0.75)


```


In total the Velmeshev et al contains `r length(unique(colnames(meta)))` metadata columns. Here is the metadata summary:  

```{r}
res <- profile_df(meta)

res$numeric_summary %>% DT::datatable()
```


```{r}
res$categorical_tables %>%
  do.call(rbind,.) %>%
  DT::datatable()
```

```{r, fig.width = 10, fig.height = 14}
wrap_plots(res$categorical_plots, ncol = 3)
```

```{r, fig.width = 10, fig.height = 14}
wrap_plots(res$numeric_plots, ncol = 3)
```



### Export
```{r}

write.csv(dat, file = file.path(paths$single_nucleus$formatted$velmeshev$dir, paths$single_nucleus$formatted$velmeshev$expr), quote = FALSE)

write.csv(meta, file = file.path(paths$single_nucleus$formatted$velmeshev$dir, paths$single_nucleus$formatted$velmeshev$metadata), quote = FALSE, row.names = FALSE)

```


```{r}
rm(meta)
rm(dat)
rm(vel.annot)
rm(VL)
```




## Cameron et al

Sn-RNA-seq data were obtained from the study by Cameron et al., which profiled transcriptional programs in the prenatal human brain. This dataset comprises snRNA-seq profiles generated from second-trimester fetal brain tissue sampled across five regions of relevance to neurodevelopmental disorders: frontal cortex (FC), ganglionic eminence (GE), hippocampus, thalamus, and cerebellum.


Processed expression matrices were downloaded from Figshare at:
https://figshare.com/articles/dataset/11629311.

Expression files were provided separately for each brain region and were analysed on a per-region basis.

The original study is described in Cameron et al., Biological Psychiatry (2022).

```{r}


cam.annot <- read.csv(here::here("data/annotations/Cameron-metadata-annot.csv"),
                      header=TRUE,
                      check.names=FALSE)
cam.dir <- paths$single_nucleus$raw$cameron

strc <- c("Thal", "Cer", "FC", "Hipp", "GE")


region_by_ds <- c(
  Thal = "Forebrain",
  Cer  = "Hindbrain",
  FC   = "Forebrain",
  Hipp = "Forebrain",
  GE   = "Forebrain"
)

structure_by_ds <- c(
  Thal = "DTH",
  Cer  = "CB",
  FC   = "DLPFC",
  Hipp = "HIP",
  GE   = "GE"
)


e <- setNames(strc, strc) %>%
  map(\(s) {
    file <- file.path(cam.dir, paste0("cameron_2021_snRNAseq_", s, "_raw_count_gEX_matrix.txt"))
    fread(file) %>%
      as.data.frame() %>%
      column_to_rownames("V1")
  })

md <- setNames(strc, strc) %>%
  map(\(s) {
    file <- file.path(cam.dir, paste0("cameron_2021_snRNAseq_", s, "_metadata.txt"))
    read.table(file, header = TRUE, stringsAsFactors = FALSE)
  })


exp_mat <- list()
md_ful  <- list()

for (f in strc) {

  name <- f
  meta <- md[[f]]
  exp_dat <- e[[f]]

  message("Now pre-processing ", name)

  ser <- CreateSeuratObject(
    counts = exp_dat,
    min.cells = round(ncol(exp_dat) / 100),
    min.features = min.features,
    project = name
  )

  # Align metadata rows to Seurat cells
  m1 <- match(colnames(ser), meta$cells)
  ser$orig.celltype <- meta$cellIDs[m1]

  # Keep only cells with a non-empty cell type
  keep_cells <- colnames(ser)[!is.na(ser$orig.celltype) & ser$orig.celltype != ""]
  ser <- subset(ser, cells = keep_cells)

  # preprocess
  max.depth <- get.max.depth(ser)
  ser <- preprocess.fun(ser, max.depth = max.depth)

  # QC vars
  nCount = ser$nCount_RNA %>%
      reshape2::melt() %>% set_colnames("nCount") %>%
      rownames_to_column("CellID")

  nFeature = ser$nFeature_RNA %>%
      reshape2::melt() %>% set_colnames("nFeature") %>%
      rownames_to_column("CellID")

  # CPM matrix
  dat <- GetAssayData(object = ser, assay = "RNA", slot = "counts") %>% as.data.frame()
  dat <- make.cpm(dat)
  dat <- rownames_to_column(dat, var = "Symbol")
  rownames(dat) <- seq_len(nrow(dat))

  # Re-align meta
  m2 <- match(colnames(ser), meta$cells)
  meta2 <- meta[m2, , drop = FALSE]

  # Metadata standardisation + cell type grouping (all in mutate; no helper funcs)
  meta2 <- meta2 %>%
    dplyr::rename(CellID = cells, `Cell Type` = cellIDs, DonorID = sample) %>%
    mutate(
      Regions = unname(region_by_ds[name]),
      `Structure Acroynm` = unname(structure_by_ds[name]),
      `Age Interval` = "13-15pcw",

      `Major Cell Type` = case_when(
        str_detect(`Cell Type`, "Ex")      ~ "Excitatory Neurons",
        str_detect(`Cell Type`, "InN")     ~ "Inhibitory Neurons",
        str_detect(`Cell Type`, "MG")      ~ "Microglia",
        str_detect(`Cell Type`, "OPC")     ~ "OPC",
        str_detect(`Cell Type`, "CycPro")  ~ "Cycling progenitor cell",
        str_detect(`Cell Type`, "IP")      ~ "Intermediate progenitor",
        str_detect(`Cell Type`, "RG")      ~ "Radial glia",
        str_detect(`Cell Type`, "Endo")    ~ "Endothelial",
        str_detect(`Cell Type`, "CR")      ~ "Cajal-Retzius cell",
        TRUE                            ~ "Undefined Class Neurons"
      ),

      `Cell Type Group` = case_when(
        str_detect(`Major Cell Type`, "Neurons") ~ "Neurons",
        `Major Cell Type` %in% c("Microglia", "OPC") ~ "Glia",
        `Major Cell Type` == "Endothelial" ~ "Vasculature",
        TRUE ~ "Developing"
      )
    ) %>%
      left_join(nCount, by = "CellID") %>%
      left_join(nFeature, by = "CellID")

  md_ful[[name]]  <- meta2
  exp_mat[[name]] <- dat
}

md.cameron <- bind_rows(md_ful, .id = "Structure") %>%
  mutate(SampleID = DonorID) %>%
  mutate(DonorID  = sub("_.*$", "", DonorID)) %>%
  mutate(DonorID = paste0(DonorID, "_donorID"))




joined <- exp_mat %>%
  reduce(full_join, by = "Symbol") %>%
  mutate(across(-Symbol, ~replace_na(.x, 0)))
```



After harmonization of metadata, a series of exploratory analyses, including correlation analysis was performed to remove redundant or non-informative metadata. 

For categorical metadata, a table is shown below which total number of unique values in each category.

```{r}
cat_tbl <- cat_summary(md.cameron , max_show = 50)


cat_tbl %>% as.data.frame() %>% DT::datatable()
```


```{r}
seq_cols <- get_seq_metrics_cols(cam.annot)
seq_cols <- intersect(seq_cols, colnames(md.cameron))
```

```{r}
res <- select_seq_metrics(
  md = md.cameron,
  seq_cols = seq_cols,
  corr_thresh = 0.9,
  missing_thresh = 0.5,
  min_unique = 2,
  prefer = "less_missing"
)
res$summary %>% DT::datatable()

```

```{r fig.cap="Correlation plot showing of correlation of sequencing metrics in Cameron et al"}
md.clean = md.cameron %>%
  dplyr::select(contains(seq_cols))
#md.clean = md.clean %>% select_if(~ !any(is.na(.)))
md.clean = md.clean[vapply(md.clean, function(x) length(unique(x)) > 1, logical(1L))]

M = cor(data.matrix(md.clean), use = "complete.obs")
corrplot(M, order='hclust',
         method = "circle",
         number.cex = .50,
         tl.cex=0.75)


```


In total the Cameron et al contains `r length(unique(colnames(md.cameron)))` metadata columns. Here is the metadata summary:  

```{r}
res <- profile_df(md.cameron)

res$numeric_summary %>% DT::datatable()
```


```{r}
res$categorical_tables %>%
  do.call(rbind,.) %>%
  DT::datatable()
```

```{r, fig.width = 10, fig.height = 14}
wrap_plots(res$categorical_plots, ncol = 3)
```

```{r, fig.width = 10, fig.height = 8}
wrap_plots(res$numeric_plots, ncol = 3)
```

### Export

```{r}

md.cameron %>% 
  dplyr::select(-c(CellID, `Cell Type Group`)) %>%
  DT::datatable()


md.cameron %<>% 
  dplyr::select(-c(CellID, `Cell Type Group`)) 
```


```{r}

write.csv(joined, file = file.path(paths$single_nucleus$formatted$cameron$dir, paths$single_nucleus$formatted$cameron$expr), quote = FALSE)

write.csv(md.cameron, file = file.path(paths$single_nucleus$formatted$cameron$dir, paths$single_nucleus$formatted$cameron$metadata), quote = FALSE, row.names = FALSE)

#write.csv(md.cameron, file = file.path(here::here("output/metadata"), "Cameron-metadata.csv"), row.names = FALSE)
```






